---
title: "9. Dynamic structural equation models"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{9. Dynamic structural equation models}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

`Rceattle` is integrated with the [`dsem`](https://CRAN.R-project.org/package=dsem) package via the `dsem` argument to `fit_mod()`. A dynamic structural equation model (DSEM) lets you place explicit structural relationships on recruitment deviations — including links to environmental covariates and cross-species linkages — instead of treating them as independent random effects.

> **Requires `dsem` 2.0.1.** DSEM models call `dsem::dsem()` internally, and the bundled C++ (`src/TMB/dsem.hpp`) is matched to `dsem` version 2.0.1. Other versions emit an incompatible data layout, so `fit_mod()` stops with an error if a different `dsem` is installed. Install the matching version with `remotes::install_version("dsem", "2.0.1")`.

This vignette covers:

- Specifying a SEM with `build_DSEM()`
- Linking environmental covariates to recruitment in a single-species model
- Linking recruitment across species in a multi-species model
- Inspecting the fitted DSEM object

## `build_DSEM()` arguments

`build_DSEM()` produces the list of DSEM settings consumed by `fit_mod()`. Its main arguments:

- `sem` — text specification of the SEM. One row per link with the form `from -> to, lag, param_name, start_value` (or `<->` for a variance/covariance term). All variables must appear as columns in `data_list$env_data`, and recruitment deviations must be named `recdevs1`, `recdevs2`, … one per species. If no start value is given, that term is **not** estimated. The default (when `sem = NULL`) is IID recruitment deviates.
- `family` — measurement-error family for each variable: `"fixed"`, `"normal"`, `"gamma"`, `"bernoulli"`, or `"poisson"`. A single string is recycled across all variables.
- `all_vars` — if `TRUE`, all variables in `env_data` are included in the DSEM likelihood (with observation error). This enables model comparison across different SEM specifications. Default `FALSE`.
- `estimate_projection` — if `FALSE` (default), latent states for the projection period are turned off, simplifying estimation.

See `?build_DSEM` and `dsem::make_dsem_ram` for full details on the SEM grammar.

## Setup

```{r}
library(Rceattle)
library(dplyr)
```

## Single-species: GOA pollock

We start with single-species GOA pollock and add a bottom-temperature covariate to `env_data`. Scaling the covariate makes the SEM coefficients comparable across links.

```{r}
data("GOApollock")
GOApollock$projyr <- 2020   # Shorten projection horizon for quicker estimation
GOApollock$env_data <- GOApollock$env_data %>%
  dplyr::mutate(ScaledBT = scale(BTempC))
```

### Model 1: IID recruitment (no DSEM)

With `dsem = NULL`, recruitment deviations are treated as IID random effects — the standard `Rceattle` behaviour.

```{r}
model1 <- Rceattle::fit_mod(
  data_list    = GOApollock,
  inits        = NULL,
  file         = NULL,
  estimateMode = 0,
  random_rec   = TRUE,
  msmMode      = 0,
  phase        = FALSE,
  initMode     = 2,
  verbose      = 1)

summary(model1)
```

### Model 2: bottom temperature → recruitment

The SEM below specifies (i) a one-year AR(1) process for scaled bottom temperature, (ii) a one-year-lagged effect of temperature on recruitment, and (iii) the recruitment-deviation variance. Phasing (`phase = TRUE`) helps the model reach a sensible starting point before the SEM parameters are turned on.

```{r}
sem_iid <- "
  # link, lag, param_name, start_value

  # internal data relation
  ScaledBT -> ScaledBT,  1, AR_BT,    0
  ScaledBT -> recdevs1,  1, BT_to_R,  0
  recdevs1 <-> recdevs1, 0, sigmaR1,  1
"

model2 <- Rceattle::fit_mod(
  data_list    = GOApollock,
  inits        = NULL,
  file         = NULL,
  estimateMode = 0,
  dsem         = build_DSEM(sem    = sem_iid,
                            family = "normal"),
  random_rec   = TRUE,
  msmMode      = 0,
  phase        = TRUE,
  initMode     = 2,
  verbose      = 1)

summary(model2)
```

### Compare the two fits

```{r}
mod_list  <- list(model1, model2)
mod_names <- c("Single-species IID", "Single-species + BT")

plot_biomass(Rceattle = mod_list, model_names = mod_names)
plot_recruitment(Rceattle = mod_list, model_names = mod_names)
```

### Inspect the DSEM object

The fitted model carries the underlying `dsem` object, which can be queried for SEM structure and parameter estimates.

```{r}
class(model2$dsem)

# SEM specification stored on the model
print(model2$data_list$dsem_settings$sem)

# Parameter estimates aligned with the SEM rows
summ <- summary(model2$dsem)
summ$Estimate <- model2$estimated_params$beta_z
knitr::kable(summ)
```

## Multi-species: EBS three-species model

`recdevsN` indexes the recruitment deviations of species `N`. In a multi-species model, this lets you link recruitment processes across species — e.g. the recruitment of one species predicting the recruitment of another in the same year.

```{r}
data("BS2017SS")
BS2017SS$projyr <- 2020
```

### Model 3: three-species IID recruitment

```{r}
model3 <- Rceattle::fit_mod(
  data_list    = BS2017SS,
  inits        = NULL,
  file         = NULL,
  estimateMode = 0,
  random_rec   = TRUE,
  msmMode      = 0,
  phase        = TRUE,
  verbose      = 1)

summary(model3)
```

### Model 4: cod and ATF recruitment → pollock recruitment

Species in `BS2017SS` are ordered pollock (1), Pacific cod (2), arrowtooth flounder (3). The SEM below keeps each species' own variance and adds two same-year cross-species links into pollock recruitment.

```{r}
sem_3spp <- "
  # link, lag, param_name, start_value
  recdevs1 <-> recdevs1, 0, sigmaR1,  1
  recdevs2 <-> recdevs2, 0, sigmaR2,  1
  recdevs3 <-> recdevs3, 0, sigmaR3,  1
  recdevs2 -> recdevs1, 0, R2_to_R1, 0
  recdevs3 -> recdevs1, 0, R3_to_R1, 0
"

model4 <- Rceattle::fit_mod(
  data_list    = BS2017SS,
  inits        = NULL,
  file         = NULL,
  estimateMode = 0,
  random_rec   = TRUE,
  dsem         = build_DSEM(sem    = sem_3spp,
                            family = "normal"),
  msmMode      = 0,
  phase        = TRUE,
  verbose      = 1)

summary(model4)
```

### Compare the two fits

```{r}
mod_list  <- list(model3, model4)
mod_names <- c("Three-species IID", "Three-species linked")

plot_biomass(Rceattle = mod_list, model_names = mod_names)
plot_recruitment(Rceattle = mod_list, model_names = mod_names)
```

## Tips

- **Phasing.** DSEM-linked models can be sensitive to starting values. Setting `phase = TRUE` fits the assessment in stages and typically reaches a better starting point before the SEM parameters are turned on.
- **Estimating vs. fixing parameters.** Any SEM term written *without* a `start_value` is treated as fixed and is not estimated. Provide a start value to estimate the term.
- **Comparing SEM specifications.** To compare alternative SEMs by likelihood, set `all_vars = TRUE` so every variable in `env_data` contributes to the DSEM likelihood under each specification.
- **Projections.** Latent DSEM states are not estimated for projection years by default. Set `estimate_projection = TRUE` if you need them — this adds parameters and can complicate estimation.
