---
title: "Environmental linkages: a formula-driven, long-format API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Environmental linkages: a formula-driven, long-format API}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

Environmental drivers (temperature, prey density, climate indices)
enter stock assessments through several distinct process modules:
recruitment, natural mortality, growth, catchability, selectivity.
Rceattle exposes a **long-format linkage table** where each row
corresponds to exactly one estimated coefficient that ties an
environmental term (or any user-specified design column) to a
process parameter. Users describe the linkages they want with
`linkage_spec()` -- a formula-driven helper -- and pass the result
into the relevant `build_*()` function. Internally, `fit_mod()`
pools every spec into a single `data.frame` with one row per
`beta`, paired with a shared design matrix `X`. The TMB template
iterates the table once, accumulates per-process offsets, and adds
them to the underlying linear predictors.

This vignette covers the user-facing API, the formula syntax it
supports, prior and bound options, and the per-species variations.
The linkage path is wired for **growth** (von Bertalanffy and
Richards) and **natural mortality** (`log_M`) today; recruitment is
next on the same pipeline.

## A first example: temperature on `K`

Suppose we want sex- and species-specific von Bertalanffy growth, with
the von Bertalanffy `K` parameter responding to a temperature index by
species:

```{r}
library(Rceattle)
data(BS2017SS)

# 1. Attach an env_data table to the data list. Must include `Year`
#    plus whatever covariates show up in your formulas. Here `temp` is
#    a placeholder; in real use it would be a standardised SST series,
#    PDO index, etc.
yrs <- BS2017SS$styr:BS2017SS$endyr
BS2017SS$env_data <- data.frame(
  Year = yrs,
  temp = scale(seq_along(yrs))[, 1]
)

# 2. Build a growth specification with one linkage per parameter that
#    you want to vary. linkage_spec() captures (formula, by, link,
#    init, bounds, priors) for one process parameter.
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = linkage_spec(
      formula = ~ temp,             # 2 columns: (Intercept) and `temp`
      by      = ~ species + sex,    # one beta per (species, sex, column)
      link    = "log",              # K lives on the log scale already
      priors  = list(temp = normal(0, 1))
    )
  )
)

# 3. Fit as usual.
fit <- fit_mod(
  data_list   = BS2017SS,
  growthFun   = growth_spec,
  estimateMode = 0,
  msmMode     = 0,
  fit_control = fit_control(verbose = 0)
)
```

The pooled linkage table is stored on the fitted model:

```{r}
fit$data_list$linkage_table
#>      process    param species sex age_bin X_col link  init prior_family ...
#> 1     growth    log_K       1   1      NA     1 log   0.0         none ...
#> 2     growth    log_K       1   2      NA     1 log   0.0         none ...
#> 3     growth    log_K       2   1      NA     1 log   0.0         none ...
#> 4     growth    log_K       2   2      NA     1 log   0.0         none ...
#> 5     growth    log_K       1   1      NA     2 log   0.0       normal ...
#> 6     growth    log_K       1   2      NA     2 log   0.0       normal ...
#> 7     growth    log_K       2   1      NA     2 log   0.0       normal ...
#> 8     growth    log_K       2   2      NA     2 log   0.0       normal ...
```

For three species × two sexes × `~ temp` (intercept + slope) you get
12 rows: an intercept beta and a temp-slope beta per (species, sex)
combination. The intercept rows aren't a duplicate of `ln_growth_pars`
-- they are just zero-valued columns of the global design matrix and
typically stay at their initial value (you can use the linkage table
solely for slopes if you prefer).

The shared design matrix is stored alongside:

```{r}
head(fit$data_list$linkage_X)
#>    (Intercept)        temp
#> 1            1  -1.5933919
#> 2            1  -1.4658285
#> 3            1  -1.3382651
#> ...
```

## The `linkage_spec()` API

```r
linkage_spec(
  formula,
  param     = NULL,
  by        = NULL,
  species   = NULL,
  link      = c("identity", "log", "logit"),
  init      = NULL,
  bounds    = NULL,
  priors    = NULL,
  re_group  = NA_character_,
  est_phase = 1L
)
```

* **`formula`** -- a one-sided R formula whose RHS describes the
  linear predictor for the target parameter. Anything `model.matrix()`
  understands works: `~ 1`, `~ temp`, `~ temp + PDO`, `~ temp:PDO`,
  `~ poly(temp, 2)`, factor predictors, etc.
* **`param`** -- the target parameter name (e.g. `"log_K"`). When the
  spec is registered inside a `build_*()` linkages list keyed by
  parameter name, `param` is filled in automatically from the list
  key.
* **`by`** -- a one-sided formula naming the stratifying factors that
  should each get their own coefficients. Allowed names are
  `species`, `sex`, and `age_bin`. `by = ~species` produces one
  coefficient set per species; `by = ~species + sex` produces one
  per (species, sex) cell. `NULL` means a single coefficient applies
  to every species/sex.
* **`species`** -- optional integer vector of 1-based species ids
  this spec applies to. `NULL` (default) means every species in the
  model. Use this to register *different formulas for different
  species* under the same target parameter -- see *Per-species
  formulas* below.
* **`link`** -- the link function applied to the predictor when the
  offset is added back to the underlying parameter. `"identity"` and
  `"log"` are equivalent for parameters already on a log scale (which
  is currently true for all growth parameters). `"logit"` is reserved
  for future use (e.g. priors on steepness in `(0, 1)`).
* **`init`**, **`bounds`** -- optional named lists keyed by
  design-matrix column name. `init = list(temp = 0.05)` sets the temp
  slope's initial value to 0.05; `bounds = list(temp = c(-2, 2))`
  constrains it during optimisation.
* **`priors`** -- a named list of prior objects keyed by design-matrix
  column. Each entry is either a single `Rceattle_prior` (applied to
  every species/sex row that uses that column) or a named list of
  priors keyed by species id (one prior per species). Inside
  `linkage_spec()` (and only there) the unprefixed short-names
  `normal()`, `lognormal()`, `gamma()`, and `beta()` resolve to the
  corresponding `prior_*` constructors via a private data mask --
  see *Priors* below.
* **`re_group`** -- random-effects grouping label (currently a
  placeholder).
* **`est_phase`** -- estimation phase ordinal. `0` fixes the
  coefficient at its initial value; `1+` makes it estimable. Phased
  fitting through `fit_control(phase = TRUE)` honors the values you
  pick.

## Priors

Priors are first-class. Five families are available out of the box:

```r
prior_normal(mean, sd)
prior_lognormal(meanlog, sdlog)
prior_gamma(shape, rate)
prior_beta(shape1, shape2)
# Family "none" is implicit -- no prior contribution.
```

Inside `linkage_spec(priors = ...)` the same constructors are
available **without the `prior_` prefix** so user code stays close to
mathematical notation:

```{r}
linkage_spec(
  formula = ~ temp + PDO,
  by      = ~ species,
  priors  = list(
    temp = normal(0, 1),     # informative prior, ~95% in (-2, 2)
    PDO  = normal(0, 5)      # weak prior
  )
)
```

The unprefixed forms are **scoped** to the `priors = ...` argument and
do not mask `base::gamma()` or `base::beta()` at the package level.
Outside that argument you can still call the math functions directly:

```{r}
gamma(5)       # 24, the math gamma function
beta(2, 3)     # 1/12, the math beta function

linkage_spec(
  formula = ~ temp,
  priors  = list(temp = gamma(2, 3))   # interpreted as prior_gamma(shape=2, rate=3)
)
```

For programmatic prior assembly (e.g. species-specific priors built in
a loop) call the `prior_*` constructors directly:

```{r}
priors <- lapply(species_specific_sds, function(s) prior_normal(0, s))
names(priors) <- c("temp", "PDO", "salinity")
linkage_spec(formula = ~ temp + PDO + salinity, priors = priors)
```

Each prior's negative log-density contributes to slot 19 of the
joint NLL (`fit$quantities$jnll_comp[20, ]` in R, since R is
1-indexed):

```{r}
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## Putting a prior on the parameter itself

Sometimes you don't want to estimate an environmental slope -- you
just want a Bayesian-style informative prior on `log_K`, `log_L1`,
`log_Linf`, or (Richards) `log_m`. Two patterns work:

**Pattern A (recommended): intercept-only linkage with the underlying
`ln_growth_pars` mapped out.** The cpp computes
`growth_parameters = exp(ln_growth_pars + ln_growth_par_devs +
linkage_offset)`. If you fix `ln_growth_pars` to `0`, the linkage
intercept becomes the entire `log_K`, and a prior on the intercept is
a prior on `log_K` itself.

```{r}
# Single species, von Bertalanffy, with N(log(0.3), 0.1) prior on log_K.
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = linkage_spec(
      formula = ~ 1,                                  # intercept only
      by      = ~ species + sex,
      init    = list(`(Intercept)` = log(0.3)),       # K starting at 0.3
      priors  = list(`(Intercept)` = normal(log(0.3), 0.1))
    )
  )
)

# Build the parameter list, then fix ln_growth_pars[, , 1] (= log_K)
# at zero so the linkage intercept owns the whole log_K. (Repeat for
# log_L1 / log_Linf / log_m as needed.)
inits <- build_params(BS2017SS)
inits$ln_growth_pars[, , 1] <- 0       # K column

# Build the map and NA-out the same slice.
mp <- build_map(BS2017SS, inits)
mp$mapList$ln_growth_pars[, , 1] <- NA
mp$mapFactor <- lapply(mp$mapList, factor)

fit <- fit_mod(
  data_list = BS2017SS,
  inits     = inits,
  map       = mp,
  growthFun = growth_spec,
  estimateMode = 0,
  msmMode   = 0
)
```

**Pattern B (looser): leave `ln_growth_pars` free and put the prior
on the linkage intercept.** This adds `-dnorm(beta_intercept,
log(0.3), 0.1, log = TRUE)` to the NLL but the *underlying* `log_K`
remains `ln_growth_pars + beta_intercept`, which is non-identifiable
without further constraints. The two parameters can trade off in
optimization. Pattern B is occasionally useful as a regularizer
("the linkage intercept should be small") but it does **not** pin
`log_K` itself.

In both patterns the prior NLL contribution shows up in
`fit$quantities$jnll_comp["Linkage-table priors", ]` and you can
inspect the realised value via the row in `fit$data_list$linkage_table`
whose `prior_family != "none"`.

A direct prior on `ln_growth_pars` (analogous to `M_prior` /
`M_prior_sd` for natural mortality) is on the post-release roadmap;
until that lands, Pattern A is the recommended workaround.

## Species- and (species, sex)-specific priors

Each entry under `priors = list(...)` may be:

* a single prior -- applies to every species/sex row that uses
  that design column,
* a **named list of priors keyed by species id** -- one prior per
  species, or
* a **two-level nested list keyed first by species, then by sex
  id** -- one prior per (species, sex).

```{r}
# Species-only keying
linkage_spec(
  formula = ~ temp,
  by      = ~ species,
  priors  = list(
    temp = list(
      `1` = normal(0, 0.1),    # tight prior for species 1
      `2` = normal(0, 0.5),    # looser for species 2
      `3` = normal(0, 1.0)     # weakest for species 3
      # any species not listed gets no prior
    )
  )
)
```

For sexually dimorphic stocks, the same `temp` column can take
different priors per (species, sex). At each level the value can be
either an `Rceattle_prior` (applies to all sub-strata) or a list
that drills down further:

```{r}
linkage_spec(
  formula = ~ temp,
  by      = ~ species + sex,
  priors  = list(
    temp = list(
      # Species 1: different tightness per sex.
      `1` = list(`1` = normal(0, 0.1),    # sp 1, sex 1 (e.g. females)
                 `2` = normal(0, 0.2)),   # sp 1, sex 2 (e.g. males)
      # Species 2: scalar applies to every sex.
      `2` = normal(0, 0.5)
      # Any (species, sex) combination not listed gets no prior.
    )
  )
)
```

The same syntax works for any prior family. This is the
recommended pattern when a covariate has different prior strength
across stocks (e.g. when one species has prior literature on its
temp sensitivity but another doesn't), or when within a species
the sexes are expected to respond differently. Built
programmatically, it's just:

```{r}
sd_by_species <- c(`1` = 0.1, `2` = 0.5, `3` = 1.0)
priors_per_sp <- lapply(sd_by_species, function(s) prior_normal(0, s))

linkage_spec(
  formula = ~ temp,
  by      = ~ species,
  priors  = list(temp = priors_per_sp)
)
```

Validation: at each level the value must be either an
`Rceattle_prior` *or* a named list of priors keyed by the next
stratum. Non-prior values (numeric vectors, mismatched classes)
are caught at `linkage_spec()` time with a pointer to the offending
(column, species[, sex]) path.

## Per-species formulas

If species have qualitatively different env relationships -- e.g.
species 1 responds only to temperature, species 2 also to PDO --
register *multiple specs against the same parameter* and use the
`species` argument on each to scope it to a subset of stocks:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = list(
      linkage_spec(
        formula = ~ temp,
        by      = ~ species,
        species = 1L,                                    # only sp 1
        priors  = list(temp = normal(0, 0.5))
      ),
      linkage_spec(
        formula = ~ temp + PDO,
        by      = ~ species,
        species = 2L,                                    # only sp 2
        priors  = list(temp = normal(0, 1),
                       PDO  = normal(0, 1))
      )
    )
  )
)
```

Inside `pool_linkages()` each spec is materialized independently
and the resulting rows are concatenated into a single linkage table.
The shared design matrix unions all RHS terms across specs, so
species 1's rows reference only `(Intercept)` and `temp`, while
species 2's rows can additionally reference `PDO` -- there's no
duplication of the temp column.

`build_growth()` accepts both forms under each parameter key:

```r
linkages = list(log_K = single_spec)             # one formula for all sp
linkages = list(log_K = list(spec_a, spec_b))    # different formulas per sp
```

Inspecting the result, `fit$linkages$log_K` is either an
`Rceattle_linkage_spec` (single form) or a `list` of them
(multi-spec form), and `fit$data_list$linkage_table` shows the
materialized rows with `species` and `X_col` correctly scoped.

This pattern composes with species-specific *priors* -- each spec
carries its own priors, and within a spec the priors can themselves
be species-keyed if you want even finer control:

```{r}
linkage_spec(
  formula = ~ temp,
  by      = ~ species + sex,
  species = c(1L, 2L),
  priors  = list(
    temp = list(
      `1` = normal(0, 0.3),
      `2` = normal(0, 0.7)
    )
  )
)
```

## Non-linear and basis-expansion formulas

The formula passed to `linkage_spec()` is fed straight to
[`stats::model.matrix()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/model.matrix.html),
so any expression `model.matrix()` understands works -- polynomial
bases, splines, interactions, and arithmetic transformations.

**Polynomials.** Use `poly()` for orthogonal polynomial bases (the
default; numerically well-conditioned), or `poly(..., raw = TRUE)`
for raw monomials. Each basis order shows up as a separate design
column with its own coefficient:

```{r}
# 4th-order orthogonal polynomial in temp, by species.
linkage_spec(
  formula = ~ poly(temp, 4),
  by      = ~ species
)
# design columns produced:
#   (Intercept), poly(temp, 4)1, poly(temp, 4)2,
#   poly(temp, 4)3, poly(temp, 4)4
```

`poly()` orthogonalises the basis against the rows of `env_data`,
so the columns are zero-mean over the model years and the linear
slope, quadratic curvature, etc. are estimated independently --
generally the right choice if you don't have a specific scientific
reason to prefer raw monomials.

**Arbitrary arithmetic.** Use `I()` to insert a literal expression
(otherwise `temp^2` is interpreted as a formula crossing operator
and reduces to `temp`):

```{r}
# Linear + raw quadratic in temp.
linkage_spec(
  formula = ~ temp + I(temp^2),
  by      = ~ species,
  priors  = list(`I(temp^2)` = normal(0, 0.1))   # tight prior on curvature
)
```

The design column name is the deparsed call (`I(temp^2)`), so when
you key `init`, `bounds`, or `priors` against it you have to use
the same string -- backticks let you write it as a name.

**Splines.** Natural cubic splines via `splines::ns()` work
out-of-the-box and are often a better choice than high-order
polynomials when you want flexibility without strong tail behaviour:

```{r}
linkage_spec(
  formula = ~ splines::ns(temp, df = 4),
  by      = ~ species
)
```

**Interactions.** `temp:PDO` adds a single product column;
`temp * PDO` is shorthand for `temp + PDO + temp:PDO`:

```{r}
linkage_spec(
  formula = ~ temp * PDO,
  by      = ~ species,
  priors  = list(`temp:PDO` = normal(0, 0.5))
)
```

**Mixing forms across species.** Combined with the per-species
formula machinery, you can give one species a polynomial response
and another a linear one without duplicating columns:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = list(
      linkage_spec(formula = ~ poly(temp, 3),
                   by = ~ species, species = 1L),
      linkage_spec(formula = ~ temp,
                   by = ~ species, species = 2L)
    )
  )
)
```

`pool_linkages()` unions the design columns by name so the global
matrix `X` contains both `temp` and the three `poly(temp, 3)*`
columns; species 1's rows reference the polynomial columns, species
2's only the linear `temp`.

A practical caution: same-named columns must be numerically
identical across specs that use them (this is enforced --
`pool_linkages()` errors loudly on mismatches). With deterministic
formulas evaluated against the same `env_data` this is automatic;
it can only fail if you've built specs against different
`env_data` frames.

## What happens internally

When you call `fit_mod()` with linkages:

1. **`build_growth(fun, linkages)`** returns a list whose
   `linkages` element is a named list of `Rceattle_linkage_spec`
   objects (or, for per-species formulas, lists of them).

2. **`fit_mod()`** stashes `growth_fun` / `growth_linkages` and
   `M1_linkages` on `data_list`, runs `data_check()`, then calls
   `pool_linkages(spec_groups = list(growth = ..., M = ...))`.

3. **`pool_linkages()`** materialises every spec against
   `data_list$env_data` and the species/sex/age strata, unions the
   per-spec design columns by name into a single shared design matrix
   `X`, and remaps each row's local `X_col` to the global column
   index. The resulting `Rceattle_linkage_table` and `linkage_X`
   matrix are stored on `data_list`.

4. **`encode_linkage_for_tmb()`** turns the table into the parallel
   integer/numeric vectors that the TMB template consumes
   (`linkage_process`, `linkage_param`, `linkage_species`,
   `linkage_sex`, `linkage_age_bin`, `linkage_X_col`,
   `linkage_link`, `linkage_prior_family`, `linkage_prior_p1`,
   `linkage_prior_p2`).

5. **The TMB template** iterates the linkage rows once and builds
   per-process offset tensors:
   - `growth_linkage_offset` of shape
     `[nspp, max_sex, n_growth_params, nyrs]` via
     `rceattle_apply_linkages()` in `src/TMB/linkage.hpp`, added
     to `growth_parameters` on the log scale.
   - `M_linkage_offset` of shape `[nspp, max_sex, max_age, nyrs]`
     via `rceattle_apply_M_linkages()`, added to `ln_M1` inside
     the `M1_at_age` compute.

   Both accumulators read from the same encoded table and the same
   `ln_beta_linkage` parameter vector; rows dispatch by their
   `process` column. With no linkages the offsets stay at zero so
   existing fits are identical-by-construction.

6. **Priors and bounds** flow through the same table:
   `build_bounds()` reads each row's `lower` / `upper` into
   `bounds$lower$ln_beta_linkage` / `bounds$upper$ln_beta_linkage`,
   and the cpp adds `-d<family>(beta, p1, p2, log = TRUE)` to slot
   19 of the joint NLL for every row whose `prior_family` is not
   `"none"`.

## Natural mortality

`build_M1()` supports the same `linkages` argument. The valid
parameter key is `log_M`; the offset enters additively (on the log
scale) inside the `M1_at_age` compute, so a coefficient of `0.3` on
a normalised temperature index multiplies `M1` by `exp(0.3 *
temp[yr])` each year:

```{r}
M1_spec <- build_M1(
  M1_model = "sex_age_invariant",            # or M1_model = 1 (back-compat)
  linkages = list(
    log_M = linkage_spec(
      formula = ~ temp,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.5))
    )
  )
)

fit <- fit_mod(
  data_list   = data_list,
  M1Fun       = M1_spec,
  estimateMode = 0,
  msmMode     = 0
)
```

The structural M choices (`M1_model`) and the random-effects
choices (`M1_re`) accept either integer codes or readable strings
interchangeably:

```{r}
build_M1(M1_model = "sex_age_specific", M1_re = "ar1_age")   # = 3, 4
build_M1(M1_model = 3,                  M1_re = 4)            # equivalent
```

`linkages` on M is age-aware: by default a row's `age_bin` is `NA`
and the offset broadcasts across every age. To pin an env effect to
a specific age slice, set `by = ~ species + age_bin` and supply
`strata$age_bin` when calling materialize directly (`fit_mod()`
provides this automatically). For most operational uses the default
broadcast is what you want.

The new linkage path coexists with the legacy `M1_indices = c(...)`
column-index argument for `M1_model = 4` / `5`: both add to
`ln_M1` on the log scale and either may be used. New code should
prefer `linkages` -- it composes naturally with priors, bounds, and
per-species formulas.

Inspecting the result:

```{r}
fit$data_list$linkage_table         # rows have process == "M"
fit$quantities$M_linkage_offset     # [nspp, max_sex, max_age, nyrs]
fit$quantities$jnll_comp["Linkage-table priors", ]
```

Growth and M linkages **share the same table** -- one fit can carry
linkage rows for both processes, and the underlying coefficient
vector `ln_beta_linkage` indexes both.

## A multi-process example

The same pattern composes when more than one parameter is linked:

```{r}
growth_spec <- build_growth(
  fun = "Richards",
  linkages = list(
    log_K    = linkage_spec(
      formula = ~ temp,
      by      = ~ species + sex,
      priors  = list(temp = normal(0, 1))
    ),
    log_Linf = linkage_spec(
      formula = ~ temp + PDO,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.5),
                     PDO  = normal(0, 0.5))
    ),
    log_m    = linkage_spec(
      formula = ~ 1,             # intercept only -- no env effect
      by      = ~ species
    )
  )
)
```

The pooler will:

* compute one shared `X = model.matrix(~ temp + PDO, env_data)` of
  three columns,
* emit 4 (sp × sex) + 3 (Linf cols) × 3 (sp) + 3 (m intercepts) = 16
  rows in the linkage table,
* match same-named columns (`(Intercept)`, `temp`) across specs so
  there's no duplication in `X`.

Per-process logic (vB has no `m`; only Richards does) is enforced at
build time:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(log_m = linkage_spec(~ temp))
)
#> Error: linkages$log_m is only valid with fun = 'Richards'; ...
```

Growth and M can be linked in the same fit -- each `build_*()`
output carries its own linkages list and `fit_mod()` pools them
together against a single shared design matrix:

```{r}
fit <- fit_mod(
  data_list   = data_list,
  growthFun   = build_growth(
    fun      = "vonBertalanffy",
    linkages = list(
      log_K = linkage_spec(formula = ~ temp, by = ~ species)
    )
  ),
  M1Fun       = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      log_M = linkage_spec(formula = ~ temp, by = ~ species)
    )
  ),
  estimateMode = 0,
  msmMode     = 0
)

# Linkage table contains rows for both processes:
table(fit$data_list$linkage_table$process)
#> growth      M
#>      4      4
```

## Inspecting and debugging

A few diagnostics worth knowing:

```{r}
# The materialised table -- one row per estimated coefficient.
fit$data_list$linkage_table

# The shared design matrix -- one column per unique RHS term.
fit$data_list$linkage_X

# Per-(species, sex, growth_param, year) offsets actually applied to
# growth_parameters this iteration. Useful for verifying the linkage
# fired as you intended.
str(fit$quantities$growth_linkage_offset)

# Per-row coefficient values in the same order as `linkage_table`.
fit$estimated_params$ln_beta_linkage

# The prior NLL contribution from slot 19 of jnll_comp.
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## On the roadmap

* **Recruitment** -- the same long-format pipeline will be extended
  to recruitment, so `build_srr(linkages = list(log_alpha = ...,
  log_R0 = ...))` will follow the same shape as the growth and M
  surfaces.
* **Random-effects grouping** -- the `re_group` column on the linkage
  table is stored end-to-end; the cpp will grow a hierarchical
  penalty so coefficients sharing a `re_group` value are partially
  pooled.
* **Age-dependent linkages** -- the `age_bin` stratum is consumed
  by the M accumulator (a row's `age_bin == NA` broadcasts across
  ages; specific values pin the offset to that age slice). It is
  reserved in the table schema for other processes that may
  eventually vary by age.

## See also

* `?linkage_spec`, `?build_growth`, `?build_M1`, `?prior_normal`
* `vignette("model-parameterizations", package = "Rceattle")` for the
  full set of process options.
