---
title: "Environmental linkages and priors: a formula-driven API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Environmental linkages and priors: a formula-driven API}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

Environmental drivers (temperature, prey density, climate indices)
enter stock assessments through several distinct process modules:
recruitment, natural mortality, growth, catchability, selectivity.
Rceattle exposes a **long-format linkage table** where each row
corresponds to exactly one estimated coefficient that ties an
environmental term (or any user-specified design column) to a
process parameter. Users describe the linkages they want with
`linkage_spec()` -- a formula-driven helper -- and pass the result
into the relevant `build_*()` function. Internally, `fit_mod()`
pools every spec into a single `data.frame` with one row per
`beta`, paired with a shared design matrix `X`. The TMB template
iterates the table once, accumulates per-process offsets, and adds
them to the underlying linear predictors.

This vignette covers the user-facing API, the formula syntax it
supports, prior and bound options, and the per-species variations.
The linkage path is wired for **growth** (von Bertalanffy and
Richards), **natural mortality** (`log_M1`), and **recruitment**
(`log_R0`, `log_alpha`, `log_beta`).

## A first example: temperature on `K`

Suppose we want sex- and species-specific von Bertalanffy growth, with
the von Bertalanffy `K` parameter responding to a temperature index by
species:

```{r}
library(Rceattle)
data(BS2017SS)

# 1. Attach an env_data table to the data list. Must include `Year`
#    plus whatever covariates show up in your formulas. Here `temp` is
#    a placeholder; in real use it would be a standardised SST series,
#    PDO index, etc.
yrs <- BS2017SS$styr:BS2017SS$endyr
BS2017SS$env_data <- data.frame(
  Year = yrs,
  temp = scale(seq_along(yrs))[, 1]
)

# 2. Build a growth specification with one linkage per parameter that
#    you want to vary. linkage_spec() captures (formula, by, link,
#    init, bounds, priors) for one process parameter.
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = linkage_spec(
      formula = ~ temp,             # 2 columns: (Intercept) and `temp`
      by      = ~ species + sex,    # one beta per (species, sex, column)
      link    = "log",              # K lives on the log scale already
      priors  = list(temp = normal(0, 1))
    )
  )
)

# 3. Fit as usual.
fit <- fit_mod(
  data_list   = BS2017SS,
  growthFun   = growth_spec,
  estimateMode = 3, # No optimization b/c no CAAL data
  msmMode     = 0,
  fit_control = fit_control(verbose = 0)
)
```

The pooled linkage table is stored on the fitted model:

```{r}
fit$data_list$linkage_table
#>      process    param species sex age_bin X_col link  init prior_family ...
#> 1     growth    log_K       1   1      NA     1 log   0.0         none ...
#> 2     growth    log_K       2   1      NA     1 log   0.0         none ...
#> 3     growth    log_K       1   1      NA     2 log   0.0       normal ...
#> 4     growth    log_K       2   1      NA     2 log   0.0       normal ...
```

For three species × two sexes × `~ temp` (intercept + slope) you get
12 rows: an intercept beta and a temp-slope beta per (species, sex)
combination. Whenever a base parameter (`log_K`, `log_M1`, `log_R0`,
...) appears in any linkage row, that base parameter is automatically
mapped out (held NA in the TMB map) so the linkage entirely replaces
its role -- see *Automatic base-parameter handling* below.

The shared design matrix is stored alongside:

```{r}
head(fit$data_list$linkage_X)
#>    (Intercept)        temp
#> 1            1  -1.5933919
#> 2            1  -1.4658285
#> 3            1  -1.3382651
#> ...
```

## Automatic base-parameter handling

Every base parameter that has at least one linkage row is **mapped out
of estimation automatically**. The linkage replaces the base parameter
rather than supplementing it.

The exact behavior depends on whether the formula carries an intercept:

| Formula                    | Base parameter value | Base parameter map |
|----------------------------|----------------------|--------------------|
| `~ 1` (intercept only)     | initialised to `0`   | NA (not estimated) |
| `~ temp` (intercept + slope) | initialised to `0` | NA (not estimated) |
| `~ 0 + temp` (slope only)  | left at `build_params()` default | NA (not estimated) |

The intercept-bearing rows (`~ 1` and `~ temp`) carry the parameter's
*level*: their estimated coefficient on the `(Intercept)` column is
exactly what would otherwise be the base parameter (e.g. `log_R0` for
species `s` lives on the `(Intercept)` row whose `species == s`). The
base array slot is held at `0` so the model evaluates `(Intercept) +
(slopes)` cleanly, with no hidden offset.

Slope-only formulas (`~ 0 + temp`) do *not* identify the level. The
base parameter is still mapped out (so it cannot drift), but it stays
at the `build_params()` default (e.g. `log_R0 = 9`). If you need a
non-default level, supply it via `inits`, or add an explicit intercept
to the formula.

This rule applies uniformly to every linkage target (`log_K`, `log_L1`,
`log_Linf`, `log_m`, `log_M1`, `log_R0`, `log_alpha`, `log_beta`).
Stratification (`by`) determines *which* base-parameter slots are
affected: `by = ~ species` zeros only the species in the rows;
`by = NULL` (a fully shared linkage) zeros every species; and
`by = ~ species + sex` zeros each (species, sex) combination present
in the linkage table.

A practical implication: combining `srr_fun > 0` (Beverton-Holt or
Ricker) with `linkages = list(log_alpha = ...)` automatically removes
the base `log_alpha` from estimation. Pair the linkage with priors,
bounds, or `init` if you need to constrain the level.

### Single-sex models and `by = ~ ... + sex`

If your model is single-sex (`nsex = 1` for every species in scope),
asking for `by = ~ species + sex` produces a warning: there is only
one sex level available, so the sex stratum collapses to a single
shared level. The result is equivalent to `by = ~ species`. Models
with mixed `nsex` (some single-sex, some two-sex species) emit one
sex level per species automatically -- single-sex species get one row
per parameter, two-sex species get two.

## The `linkage_spec()` API

```r
linkage_spec(
  formula,
  param     = NULL,
  by        = ~ species,
  species   = NULL,
  link      = c("identity", "log", "logit"),
  init      = NULL,
  bounds    = NULL,
  priors    = NULL,
  re_group  = NA_character_,
  est_phase = 1L
)
```

* **`formula`** -- a one-sided R formula whose RHS describes the
  linear predictor for the target parameter. Anything `model.matrix()`
  understands works: `~ 1`, `~ temp`, `~ temp + PDO`, `~ temp:PDO`,
  `~ poly(temp, 2)`, factor predictors, etc.
* **`param`** -- the target parameter name (e.g. `"log_K"`). When the
  spec is registered inside a `build_*()` linkages list keyed by
  parameter name, `param` is filled in automatically from the list
  key.
* **`by`** -- a one-sided formula naming the stratifying factors that
  should each get their own coefficients. Allowed names are
  `species`, `sex`, and `age_bin`. The default is `~species`
  (one coefficient set per species, the typical multispecies
  assessment use case). Pass `~species + sex` for per-(species, sex)
  coefficients, `~species + age_bin` for an age-binned M response,
  or `NULL` to share a single coefficient set across every
  species/sex.
* **`species`** -- optional integer vector of 1-based species ids
  this spec applies to. `NULL` (default) means every species in the
  model. Use this to register *different formulas for different
  species* under the same target parameter -- see *Per-species
  formulas* below.
* **`link`** -- the link function applied to the predictor (NOT YET IMPLEMENTED)
* **`init`**, **`bounds`** -- optional named lists keyed by
  design-matrix column name. `init = list(temp = 0.05)` sets the temp
  slope's initial value to 0.05; `bounds = list(temp = c(-2, 2))`
  constrains it during optimisation.
* **`priors`** -- a named list of prior objects keyed by design-matrix
  column. Each entry is either a single `Rceattle_prior` (applied to
  every species/sex row that uses that column) or a named list of
  priors keyed by species id (one prior per species). Inside
  `linkage_spec()` (and only there) the unprefixed short-names
  `normal()`, `lognormal()`, `gamma()`, and `beta()` resolve to the
  corresponding `prior_*` constructors via a private data mask --
  see *Priors* below.
* **`re_group`** -- random-effects grouping label (currently a
  placeholder).
* **`est_phase`** -- estimation phase ordinal. `0` fixes the
  coefficient at its initial value; `1+` makes it estimable. Phased
  fitting through `fit_control(phase = TRUE)` honors the values you
  pick.

## Priors

Priors are first-class. Five families are available out of the box:

```r
prior_normal(mean, sd)
prior_lognormal(meanlog, sdlog)
prior_gamma(shape, rate)
prior_beta(shape1, shape2)
# Family "none" is implicit -- no prior contribution.
```

Inside `linkage_spec(priors = ...)` the same constructors are
available **without the `prior_` prefix** so user code stays close to
mathematical notation:

```{r}
linkage_spec(
  formula = ~ temp + PDO,
  by      = ~ species,
  priors  = list(
    temp = normal(0, 1),     # informative prior, ~95% in (-2, 2)
    PDO  = normal(0, 5)      # weak prior
  )
)
```

The unprefixed forms are **scoped** to the `priors = ...` argument and
do not mask `base::gamma()` or `base::beta()` at the package level.
Outside that argument you can still call the math functions directly:

```{r}
gamma(5)       # 24, the math gamma function
beta(2, 3)     # 1/12, the math beta function

linkage_spec(
  formula = ~ temp,
  priors  = list(temp = gamma(2, 3))   # interpreted as prior_gamma(shape=2, rate=3)
)
```

For programmatic prior assembly (e.g. species-specific priors built in
a loop) call the `prior_*` constructors directly:

```{r}
priors <- lapply(species_specific_sds, function(s) prior_normal(0, s))
names(priors) <- c("temp", "PDO", "salinity")
linkage_spec(formula = ~ temp + PDO + salinity, priors = priors)
```

Each prior's negative log-density contributes to slot 19 of the
joint NLL (`fit$quantities$jnll_comp[20, ]` in R, since R is
1-indexed):

```{r}
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## Putting a prior on the parameter itself

Sometimes, instead of estimating how a parameter changes with
environmental covariates, you want to apply a Bayesian-style
informative prior directly to the base value of a parameter like
`log_K`, `log_R0`, or `log_M1`.

Because any intercept-bearing linkage replaces the base parameter (see
*Automatic base-parameter handling* above), an intercept-only linkage
*is* the parameter, and a prior on its `(Intercept)` row is the prior
on that parameter:

```{r}
# Example: prior on log_K for von Bertalanffy growth
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = linkage_spec(
      formula = ~ 1,                              # intercept only -- no env effects
      by      = ~ species + sex,
      init    = list(`(Intercept)` = log(0.3)),   # start at K = 0.3
      priors  = list(`(Intercept)` = normal(log(0.3), 0.1))
    )
  )
)
```

The same pattern works for any linkage target -- `log_Linf`, `log_L1`,
`log_m` (Richards), `log_M1`, `log_R0`, `log_alpha`, `log_beta`. With
recruitment, you can also key the prior by species id for species-
specific priors:

```{r}
rec_spec <- build_srr(
  srr_fun = 0,
  linkages = list(
    log_R0 = linkage_spec(
      formula = ~ 1,
      by      = ~ species,
      init    = list(`(Intercept)` = log(1000)),
      priors  = list(`(Intercept)` = list(
        `1` = normal(log(1000), 0.2),
        `2` = normal(log(1500), 0.3),
        `3` = normal(log(800),  0.25)
      ))
    )
  )
)
```

The prior contribution appears in slot 19 of the joint NLL:

```{r}
fit$quantities$jnll_comp["Linkage-table priors", ]
```

You can inspect the linked parameter values via the linkage table and
the estimated coefficient vector:

```{r}
fit$data_list$linkage_table[fit$data_list$linkage_table$param == "log_R0", ]
fit$estimated_params$ln_beta_linkage
```

The same approach works for **`log_M1`** when you want a prior on
natural mortality itself:

```{r}
# Three-species model, Lognormal(log(0.4), 0.35) prior on log_M1
# applied via the linkage intercept. The structural M1_model is
# kept at "sex_age_invariant" (= 1).
M1_spec <- build_M1(
  M1_model = "sex_age_invariant",
  linkages = list(
    log_M1 = linkage_spec(
      formula = ~ 1,
      by      = ~ species,
      init    = list(`(Intercept)` = log(0.4)),
      priors  = list(`(Intercept)` = lognormal(log(0.4), 0.35))
    )
  )
)

# Fit the model - no manual mapping needed
fit <- fit_mod(
  data_list = data_list,
  M1Fun     = M1_spec,
  estimateMode = 0
)
```

Compared to the existing `build_M1(M_prior = ..., M_prior_sd = ...,
M1_use_prior = TRUE)` mechanism, the linkage-intercept route gives
you the same prior shape but composes naturally with the rest of
the linkage table (you can keep your env-effect rows and the
intercept rows together, with priors on each). For most operational
work the existing `M_prior` mechanism is simpler and more direct;
the linkage route is most useful when you already have a table of
env-effect coefficients and want to extend the same per-row
prior/bound machinery to the parameter intercept.

## Species- and (species, sex)-specific priors

Each entry under `priors = list(...)` may be:

* a single prior -- applies to every species/sex row that uses
  that design column,
* a **named list of priors keyed by species id** -- one prior per
  species, or
* a **two-level nested list keyed first by species, then by sex
  id** -- one prior per (species, sex).

```{r}
# Species-only keying
linkage_spec(
  formula = ~ temp,
  by      = ~ species,
  priors  = list(
    temp = list(
      `1` = normal(0, 0.1),    # tight prior for species 1
      `2` = normal(0, 0.5),    # looser for species 2
      `3` = normal(0, 1.0)     # weakest for species 3
      # any species not listed gets no prior
    )
  )
)
```

For sexually dimorphic stocks, the same `temp` column can take
different priors per (species, sex). At each level the value can be
either an `Rceattle_prior` (applies to all sub-strata) or a list
that drills down further:

```{r}
linkage_spec(
  formula = ~ temp,
  by      = ~ species + sex,
  priors  = list(
    temp = list(
      # Species 1: different tightness per sex.
      `1` = list(`1` = normal(0, 0.1),    # sp 1, sex 1 (e.g. females)
                 `2` = normal(0, 0.2)),   # sp 1, sex 2 (e.g. males)
      # Species 2: scalar applies to every sex.
      `2` = normal(0, 0.5)
      # Any (species, sex) combination not listed gets no prior.
    )
  )
)
```

The same syntax works for any prior family. This is the
recommended pattern when a covariate has different prior strength
across stocks (e.g. when one species has prior literature on its
temp sensitivity but another doesn't), or when within a species
the sexes are expected to respond differently. Built
programmatically, it's just:

```{r}
sd_by_species <- c(`1` = 0.1, `2` = 0.5, `3` = 1.0)
priors_per_sp <- lapply(sd_by_species, function(s) prior_normal(0, s))

linkage_spec(
  formula = ~ temp,
  by      = ~ species,
  priors  = list(temp = priors_per_sp)
)
```

Validation: at each level the value must be either an
`Rceattle_prior` *or* a named list of priors keyed by the next
stratum. Non-prior values (numeric vectors, mismatched classes)
are caught at `linkage_spec()` time with a pointer to the offending
(column, species[, sex]) path.

## Per-species formulas

If species have qualitatively different env relationships -- e.g.
species 1 responds only to temperature, species 2 also to PDO --
register *multiple specs against the same parameter* and use the
`species` argument on each to scope it to a subset of stocks:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = list(
      linkage_spec(
        formula = ~ temp,
        by      = ~ species,
        species = 1L,                                    # only sp 1
        priors  = list(temp = normal(0, 0.5))
      ),
      linkage_spec(
        formula = ~ temp + PDO,
        by      = ~ species,
        species = 2L,                                    # only sp 2
        priors  = list(temp = normal(0, 1),
                       PDO  = normal(0, 1))
      )
    )
  )
)
```

Inside `pool_linkages()` each spec is materialized independently
and the resulting rows are concatenated into a single linkage table.
The shared design matrix unions all RHS terms across specs, so
species 1's rows reference only `(Intercept)` and `temp`, while
species 2's rows can additionally reference `PDO` -- there's no
duplication of the temp column.

`build_growth()` accepts both forms under each parameter key:

```r
linkages = list(log_K = single_spec)             # one formula for all sp
linkages = list(log_K = list(spec_a, spec_b))    # different formulas per sp
```

Inspecting the result, `fit$linkages$log_K` is either an
`Rceattle_linkage_spec` (single form) or a `list` of them
(multi-spec form), and `fit$data_list$linkage_table` shows the
materialized rows with `species` and `X_col` correctly scoped.

This pattern composes with species-specific *priors* -- each spec
carries its own priors, and within a spec the priors can themselves
be species-keyed if you want even finer control:

```{r}
linkage_spec(
  formula = ~ temp,
  by      = ~ species + sex,
  species = c(1L, 2L),
  priors  = list(
    temp = list(
      `1` = normal(0, 0.3),
      `2` = normal(0, 0.7)
    )
  )
)
```

## Non-linear and basis-expansion formulas

The formula passed to `linkage_spec()` is fed straight to
[`stats::model.matrix()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/model.matrix.html),
so any expression `model.matrix()` understands works -- polynomial
bases, splines, interactions, and arithmetic transformations.

**Polynomials.** Use `poly()` for orthogonal polynomial bases (the
default; numerically well-conditioned), or `poly(..., raw = TRUE)`
for raw monomials. Each basis order shows up as a separate design
column with its own coefficient:

```{r}
# 4th-order orthogonal polynomial in temp, by species.
linkage_spec(
  formula = ~ poly(temp, 4),
  by      = ~ species
)
# design columns produced:
#   (Intercept), poly(temp, 4)1, poly(temp, 4)2,
#   poly(temp, 4)3, poly(temp, 4)4
```

`poly()` orthogonalises the basis against the rows of `env_data`,
so the columns are zero-mean over the model years and the linear
slope, quadratic curvature, etc. are estimated independently --
generally the right choice if you don't have a specific scientific
reason to prefer raw monomials.

**Arbitrary arithmetic.** Use `I()` to insert a literal expression
(otherwise `temp^2` is interpreted as a formula crossing operator
and reduces to `temp`):

```{r}
# Linear + raw quadratic in temp.
linkage_spec(
  formula = ~ temp + I(temp^2),
  by      = ~ species,
  priors  = list(`I(temp^2)` = normal(0, 0.1))   # tight prior on curvature
)
```

The design column name is the deparsed call (`I(temp^2)`), so when
you key `init`, `bounds`, or `priors` against it you have to use
the same string -- backticks let you write it as a name.

**Splines.** Natural cubic splines via `splines::ns()` work
out-of-the-box and are often a better choice than high-order
polynomials when you want flexibility without strong tail behaviour:

```{r}
linkage_spec(
  formula = ~ splines::ns(temp, df = 4),
  by      = ~ species
)
```

**Interactions.** `temp:PDO` adds a single product column;
`temp * PDO` is shorthand for `temp + PDO + temp:PDO`:

```{r}
linkage_spec(
  formula = ~ temp * PDO,
  by      = ~ species,
  priors  = list(`temp:PDO` = normal(0, 0.5))
)
```

**Mixing forms across species.** Combined with the per-species
formula machinery, you can give one species a polynomial response
and another a linear one without duplicating columns:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = list(
      linkage_spec(formula = ~ poly(temp, 3),
                   by = ~ species, species = 1L),
      linkage_spec(formula = ~ temp,
                   by = ~ species, species = 2L)
    )
  )
)
```

`pool_linkages()` unions the design columns by name so the global
matrix `X` contains both `temp` and the three `poly(temp, 3)*`
columns; species 1's rows reference the polynomial columns, species
2's only the linear `temp`.

A practical caution: same-named columns must be numerically
identical across specs that use them (this is enforced --
`pool_linkages()` errors loudly on mismatches). With deterministic
formulas evaluated against the same `env_data` this is automatic;
it can only fail if you've built specs against different
`env_data` frames.

## What happens internally

When you call `fit_mod()` with linkages:

1. **`build_growth(fun, linkages)`** returns a list whose
   `linkages` element is a named list of `Rceattle_linkage_spec`
   objects (or, for per-species formulas, lists of them).

2. **`fit_mod()`** stashes `growth_fun` / `growth_linkages` and
   `M1_linkages` on `data_list`, runs `data_check()`, then calls
   `pool_linkages(spec_groups = list(growth = ..., M = ...))`.

3. **`pool_linkages()`** materialises every spec against
   `data_list$env_data` and the species/sex/age strata, unions the
   per-spec design columns by name into a single shared design matrix
   `X`, and remaps each row's local `X_col` to the global column
   index. The resulting `Rceattle_linkage_table` and `linkage_X`
   matrix are stored on `data_list`.

4. **`encode_linkage_for_tmb()`** turns the table into the parallel
   integer/numeric vectors that the TMB template consumes
   (`linkage_process`, `linkage_param`, `linkage_species`,
   `linkage_sex`, `linkage_age_bin`, `linkage_X_col`,
   `linkage_link`, `linkage_prior_family`, `linkage_prior_p1`,
   `linkage_prior_p2`).

5. **The TMB template** iterates the linkage rows once and builds
   per-process offset tensors:
   - `growth_linkage_offset` of shape
     `[nspp, max_sex, n_growth_params, nyrs]` via
     `rceattle_apply_linkages()` in `src/TMB/linkage.hpp`, added
     to `growth_parameters` on the log scale.
   - `M_linkage_offset` of shape `[nspp, max_sex, max_age, nyrs]`
     via `rceattle_apply_M_linkages()`, added to `ln_M1` inside
     the `M1_at_age` compute.
   - `recruitment_linkage_offset` of shape
     `[nspp, n_rec_params, nyrs]` via
     `rceattle_apply_recruitment_linkages()`, pre-added to
     `(R0, rec_pars(*, alpha), rec_pars(*, beta))` before each
     `calculate_recruitment()` call site (hindcast, BRPs, dynamic
     BRPs, projections, expected R).

   All three accumulators read from the same encoded table and the
   same `ln_beta_linkage` parameter vector; rows dispatch by their
   `process` column. With no linkages the offsets stay at zero so
   existing fits are identical-by-construction.

6. **Priors and bounds** flow through the same table:
   `build_bounds()` reads each row's `lower` / `upper` into
   `bounds$lower$ln_beta_linkage` / `bounds$upper$ln_beta_linkage`,
   and the cpp adds `-d<family>(beta, p1, p2, log = TRUE)` to slot
   19 of the joint NLL for every row whose `prior_family` is not
   `"none"`.

## Natural mortality

`build_M1()` supports the same `linkages` argument. The valid
parameter key is `log_M1`; the offset enters additively (on the log
scale) inside the `M1_at_age` compute, so a coefficient of `0.3` on
a normalised temperature index multiplies `M1` by `exp(0.3 *
temp[yr])` each year.

**Why `log_M1` and not `log_M`?** CEATTLE decomposes total natural
mortality into `M = M1 + M2`. `M1` is residual / non-predation
mortality and is a *parameter* (`ln_M1` in the cpp, with optional
random-effect deviates and a direct prior via `M_prior` /
`M_prior_sd`). `M2` is *predation mortality* and is a derived
quantity each iteration -- a function of predator abundance,
suitability, and predator ration. There is no scalar `ln_M2`
parameter to attach a linkage coefficient to. Environmental effects
on predation propagate naturally via upstream parameters: a temp
effect on predator growth (`log_K`), a temp effect on recruitment
(forthcoming), or env-driven inputs to the bioenergetics ration
calculation. The linkage system therefore exposes `log_M1` as the
single M-side target, with `M2` covered by upstream propagation.

```{r}
M1_spec <- build_M1(
  M1_model = "sex_age_invariant",            # or M1_model = 1 (back-compat)
  linkages = list(
    log_M1 = linkage_spec(
      formula = ~ temp,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.5))
    )
  )
)

fit <- fit_mod(
  data_list   = data_list,
  M1Fun       = M1_spec,
  estimateMode = 0,
  msmMode     = 0
)
```

The structural M choices (`M1_model`) and the random-effects
choices (`M1_re`) accept either integer codes or readable strings
interchangeably:

```{r}
build_M1(M1_model = "sex_age_specific", M1_re = "ar1_age")   # = 3, 4
build_M1(M1_model = 3,                  M1_re = 4)            # equivalent
```

`linkages` on M is age-aware: by default a row's `age_bin` is `NA`
and the offset broadcasts across every age. To pin an env effect to
a specific age slice, set `by = ~ species + age_bin` and supply
`strata$age_bin` when calling materialize directly (`fit_mod()`
provides this automatically). For most operational uses the default
broadcast is what you want.

The new linkage path coexists with the legacy `M1_indices = c(...)`
column-index argument for `M1_model = 4` / `5`: both add to
`ln_M1` on the log scale and either may be used. New code should
prefer `linkages` -- it composes naturally with priors, bounds, and
per-species formulas.

Inspecting the result:

```{r}
fit$data_list$linkage_table         # rows have process == "M"
fit$quantities$M_linkage_offset     # [nspp, max_sex, max_age, nyrs]
fit$quantities$jnll_comp["Linkage-table priors", ]
```

Growth and M linkages **share the same table** -- one fit can carry
linkage rows for both processes, and the underlying coefficient
vector `ln_beta_linkage` indexes both.

## Recruitment

`build_srr()` accepts the same `linkages` argument with three
allowed parameter keys:

* **`log_R0`** -- offset on `log(R0)`. Meaningful for *any*
  `srr_fun`; the offset is added to the equilibrium / mean
  recruitment on the log scale, including the first model year.
  This is the right target if you want "recruitment responds to
  temp" and aren't fitting an SRR.
* **`log_alpha`** -- offset on the alpha parameter (productivity)
  of a Beverton-Holt or Ricker SRR. Only does work when
  `srr_fun %in% c(2, 3, 4, 5)` -- the SRRs that consume alpha.
* **`log_beta`** -- offset on the beta parameter
  (density-dependence) of BH or Ricker. Same applicability as
  `log_alpha`.

```{r}
# Mean-recruitment with a temperature linkage on log_R0.
rec_spec <- build_srr(
  srr_fun  = "mean",                       # = 0
  linkages = list(
    log_R0 = linkage_spec(
      formula = ~ temp,
      priors  = list(temp = normal(0, 0.5))
    )
  )
)

# Beverton-Holt with env effects on alpha (and species-keyed
# priors so each stock has its own sensitivity).
rec_spec_bh <- build_srr(
  srr_fun  = "BevertonHolt",               # = 2
  linkages = list(
    log_alpha = linkage_spec(
      formula = ~ temp,
      priors  = list(
        temp = list(`1` = normal(0, 0.3),
                    `2` = normal(0, 0.7))
      )
    )
  )
)

fit <- fit_mod(
  data_list   = data_list,
  recFun      = rec_spec,
  estimateMode = 0,
  msmMode     = 0
)
```

The structural choice (`srr_fun`) accepts either integer codes or
readable strings interchangeably:

```{r}
build_srr(srr_fun = "mean")           # = 0
build_srr(srr_fun = "BevertonHolt")   # = 2
build_srr(srr_fun = "Ricker")         # = 4
```

Inspecting the result:

```{r}
fit$data_list$linkage_table             # rows have process == "recruitment"
fit$quantities$recruitment_linkage_offset  # [nspp, n_rec_params, nyrs]
                                           # n_rec_params = 3 (R0, alpha, beta)
```

Practical note: a linkage row's `age_bin` is ignored for
recruitment (recruitment is at age 0 by definition). The
`sex_ratio` data input partitions recruits across sexes downstream.

## A multi-process example

The same pattern composes when more than one parameter is linked:

```{r}
growth_spec <- build_growth(
  fun = "Richards",
  linkages = list(
    log_K    = linkage_spec(
      formula = ~ temp,
      by      = ~ species + sex,
      priors  = list(temp = normal(0, 1))
    ),
    log_Linf = linkage_spec(
      formula = ~ temp + PDO,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.5),
                     PDO  = normal(0, 0.5))
    ),
    log_m    = linkage_spec(
      formula = ~ 1,             # intercept only -- no env effect
      by      = ~ species
    )
  )
)
```

The pooler will:

* compute one shared `X = model.matrix(~ temp + PDO, env_data)` of
  three columns,
* emit 4 (sp × sex) + 3 (Linf cols) × 3 (sp) + 3 (m intercepts) = 16
  rows in the linkage table,
* match same-named columns (`(Intercept)`, `temp`) across specs so
  there's no duplication in `X`.

Per-process logic (vB has no `m`; only Richards does) is enforced at
build time:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(log_m = linkage_spec(~ temp))
)
#> Error: linkages$log_m is only valid with fun = 'Richards'; ...
```

Growth, M, and recruitment can all be linked in the same fit --
each `build_*()` output carries its own linkages list and
`fit_mod()` pools them against a single shared design matrix:

```{r}
fit <- fit_mod(
  data_list   = data_list,
  growthFun   = build_growth(
    fun      = "vonBertalanffy",
    linkages = list(
      log_K = linkage_spec(formula = ~ temp)
    )
  ),
  M1Fun       = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      log_M1 = linkage_spec(formula = ~ temp)
    )
  ),
  recFun      = build_srr(
    srr_fun  = "mean",
    linkages = list(
      log_R0 = linkage_spec(formula = ~ temp)
    )
  ),
  estimateMode = 0,
  msmMode     = 0
)

# Linkage table contains rows for all three processes:
table(fit$data_list$linkage_table$process)
#>      growth           M  recruitment
#>           4           4            4
```

## Inspecting and debugging

A few diagnostics worth knowing:

```{r}
# The materialised table -- one row per estimated coefficient.
fit$data_list$linkage_table

# The shared design matrix -- one column per unique RHS term.
fit$data_list$linkage_X

# Per-(species, sex, growth_param, year) offsets actually applied to
# growth_parameters this iteration. Useful for verifying the linkage
# fired as you intended.
str(fit$quantities$growth_linkage_offset)
str(fit$quantities$M_linkage_offset)
str(fit$quantities$recruitment_linkage_offset)

# Per-row coefficient values in the same order as `linkage_table`.
fit$estimated_params$ln_beta_linkage

# The prior NLL contribution from slot 19 of jnll_comp.
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## On the roadmap

* **Random-effects grouping** -- the `re_group` column on the linkage
  table is stored end-to-end; the cpp will grow a hierarchical
  penalty so coefficients sharing a `re_group` value are partially
  pooled.
* **Age-dependent linkages on processes other than M** -- the
  `age_bin` stratum is consumed by the M accumulator today (a
  row's `age_bin == NA` broadcasts across ages; specific values
  pin the offset to that age slice). The schema is reserved for
  other processes that may eventually vary by age.
* **Selectivity / catchability linkages** -- the encoder already
  reserves `process = "sel"` and `process = "q"` codes, but no
  cpp accumulator consumes them yet.
* **Legacy retirement (v4.2.0)** -- `srr_indices`, `M1_indices`,
  `srr_fun in c(1, 3, 5)`, and `M1_model in c(4, 5)` retire in
  the next minor release; see *Scheduled removal (v4.2.0)* in
  `NEWS.md` for the migration table.

## See also

* `?linkage_spec`, `?build_growth`, `?build_M1`, `?build_srr`, `?prior_normal`
* `vignette("model-parameterizations", package = "Rceattle")` for the
  full set of process options.
