---
title: "Environmental linkages and priors: a formula-driven API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Environmental linkages and priors: a formula-driven API}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

Environmental drivers (temperature, prey density, climate indices)
enter stock assessments through recruitment, natural mortality,
growth, catchability, and selectivity. Rceattle exposes a
**long-format linkage table**: one row per estimated coefficient
tying an environmental term (or any user-specified design column) to
a process parameter. Users describe linkages with `linkage_spec()` --
a formula-driven helper -- and pass the result to the relevant
`build_*()` function. `fit_mod()` pools every spec into a single
`data.frame` paired with a shared design matrix; the TMB template
iterates the table once and adds per-process offsets to the
underlying linear predictors.

The linkage path is wired for **growth** (von Bertalanffy and
Richards), **natural mortality** (`M1`), and **recruitment**
(`R0`, `alpha`, `beta`).

## A first example: temperature on `K`

Suppose we want sex- and species-specific von Bertalanffy growth, with
the von Bertalanffy `K` parameter responding to a temperature index by
species:

```{r}
library(Rceattle)
data(whamGrowthData)
plot_data(whamGrowthData)

# Use logistic selectivity for all fleets in this example.
whamGrowthData$fleet_control$Selectivity <- "Logistic"
whamGrowthData$fleet_control$Selectivity_dimension <- "Length"

# 1. Attach an env_data table to the data list. Must include `Year`
#    plus whatever covariates show up in your formulas. Here `temp` is
#    a placeholder; in real use it would be a standardised SST series,
#    PDO index, etc.
yrs <- whamGrowthData$styr:whamGrowthData$endyr
whamGrowthData$env_data <- data.frame(
  Year = yrs,
  temp = rnorm(length(yrs), 0, 0.1)
)

# 2. Build a growth specification with one linkage per parameter that
#    you want to vary. linkage_spec() captures (formula, by, link,
#    init, bounds, priors) for one process parameter.
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    K = linkage_spec(
      formula = ~ temp,             # 2 columns: (Intercept) and `temp`
      by      = ~ species + sex,    # one beta per (species, sex, column)
      priors  = list(temp = normal(0, 0.1)) # Tight prior
    )
  )
)

# 3. Fit as usual.
fit <- fit_mod(
  data_list   = whamGrowthData,
  growthFun   = growth_spec,
  estimateMode = 0, 
  msmMode     = 0,
  fit_control = fit_control(verbose = 0)
)
summary(fit)
```


## The `linkage_spec()` API

```r
linkage_spec(
  formula,
  param     = NULL,
  by        = ~ species,
  species   = NULL,
  sex       = NULL,
  link      = "log",                # or "identity"
  init      = NULL,
  bounds    = NULL,
  priors    = NULL,
  re_group  = NA_character_,
  est_phase = 1L
)
```

Key arguments (see `?linkage_spec` for the full list):

* **`formula`** -- one-sided R formula describing the linear
  predictor (anything `model.matrix()` understands: `~ 1`, `~ temp`,
  `~ temp + PDO`, `~ poly(temp, 2)`, factor predictors, etc.).
* **`param`** -- target parameter name (e.g. `"K"`). Filled in
  automatically when the spec is registered under a `linkages = list(K = ...)`
  list key.
* **`by`** -- stratifying factors (default `~species`); use
  `~species + sex` for per-(species, sex) coefficients, or `NULL`
  to share a single coefficient set.
* **`species`** / **`sex`** -- restrict a spec to specific ids.
  Use with the multi-spec form (*Per-species* / *Per-sex formulas*
  below) to give different species or sexes different formulas.
  `sex` accepts integers (`1L`, `2L`) or strings (`"Females"`/`"Males"`,
  case-insensitive).
* **`link`** -- relates the linear predictor to the natural-scale
  target. `"log"` (default) gives a multiplicative effect on natural
  `param`; `"identity"` gives an additive effect on natural `param`.
* **`init`** / **`bounds`** -- named lists keyed by design-matrix
  column (`init = list(temp = 0.05)`, `bounds = list(temp = c(-2, 2))`).
* **`priors`** -- named list keyed by design-matrix column. Each
  entry is either an `Rceattle_prior` or a list keyed by species id;
  see *Priors* below.
* **`re_group`** / **`est_phase`** -- random-effects grouping (placeholder)
  and estimation phase ordinal (`0` fixes, `1+` estimates).

## Priors

Four families are available, plus the implicit "none":

```r
prior_normal(mean, sd)
prior_lognormal(meanlog, sdlog)
prior_gamma(shape, rate)
prior_beta(shape1, shape2)
```

Inside `linkage_spec(priors = ...)` the unprefixed shorthand `normal()`,
`lognormal()`, `gamma()`, `beta()` resolves to the corresponding
`prior_*` constructors via a private data mask -- this is scoped to the
`priors` argument and does not mask `base::gamma()` / `base::beta()`
elsewhere:

```{r}
linkage_spec(
  formula = ~ temp + PDO,
  by      = ~ species,
  priors  = list(
    temp = normal(0, 1),     # informative, ~95% in (-2, 2)
    PDO  = normal(0, 5)      # weak
  )
)
```

**Prior semantics.** For `(Intercept)` rows,
Priors are evaluated on the *natural-scale* 
(e.g. `normal(0.3, 0.1)` on `K` means `dnorm(K_natural, 0.3, 0.1)`).
Use `lognormal(meanlog, sdlog)` to put a normal directly on the log
scale. For covariate rows, the prior applies
to the coefficient as-is regardless of link function.

For programmatic prior assembly (e.g. species-specific priors built in
a loop) call the `prior_*` constructors directly:

```{r}
priors <- lapply(species_specific_sds, function(s) prior_normal(0, s))
names(priors) <- c("temp", "PDO", "salinity")
linkage_spec(formula = ~ temp + PDO + salinity, priors = priors)
```

Each prior's negative log-density contributes to the `Linkage-table priors`
row of the joint NLL:

```{r}
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## Putting a prior on the parameter itself

Bayesian-style informative priors can be applied directly to the base growth
parameters (`K`, `L1`, `Linf`, `m`), natural mortality (`M1`), or recruitment parameters (`R0`, `alpha`, `beta`).

A prior attached to an `(Intercept)` row is **re-targeted to the base
parameter**. The same syntax is used for any other column.

```{r}
# Example: prior on K for von Bertalanffy growth
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    K = linkage_spec(
      formula = ~ 1,                              # intercept only -- no env effects
      by      = ~ species + sex,
      init    = list(`(Intercept)` = 0.3),   # start at K = 0.3
      # Use `lognormal(log(0.3), 0.1)` to get a normal prior on log(K).
      priors  = list(`(Intercept)` = lognormal(log(0.3), 0.1))
    )
  )
)
```

The same pattern works for any linkage target -- `Linf`, `L1`,
`m` (Richards), `M1`, `R0`, `alpha`, `beta`.

### Growth SD endpoints

The standard deviations of length-at-age at the `L1` and `Linf` anchors
(`sd_L1`, `sd_Linf`) are exposed as linkage targets too, so the
same `init` / `bounds` / `priors` contract applies. They route onto the
`growth_log_sd` parameter rather than `log_growth_pars`, and -- because
`growth_log_sd` has no year dimension in `growth.hpp` -- only
intercept-bearing formulas are honored. Slope rows on these targets
raise a warning and have no effect; slope-only formulas (`~ 0 + temp`)
error.

```{r}
# Prior on the SD at L1 anchored at exp(log(5)) = 5 length units,
# with a tight normal prior on the log scale.
growth_spec_sd <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    sd_L1   = linkage_spec(
      formula = ~ 1,
      by      = ~ species + sex,
      init    = list(`(Intercept)` = 5),
      # Prior on log-scale SD; switch to `normal(5, 0.3)`
      # for a Normal on sd_L1.
      priors  = list(`(Intercept)` = lognormal(log(5), 0.3))
    ),
    sd_Linf = linkage_spec(
      formula = ~ 1,
      by      = ~ species + sex,
      init    = list(`(Intercept)` = 15)
    )
  )
)

# After fit_mod(), the init values land on growth_log_sd[, , 1:2]
# (SD at L1 and SD at Linf, log-scale), and the priors evaluate
# against those base-parameter cells via the intercept re-targeting
# path described above.
```

You can also key the prior by species id for species-
specific priors:

```{r}
rec_spec <- build_srr(
  srr_fun = 0,
  linkages = list(
    R0 = linkage_spec(
      formula = ~ 1,
      by      = ~ species,
      init    = list(`(Intercept)` = 1000),
      # Species-keyed priors on log-scale R0.
      priors  = list(`(Intercept)` = list(
        `1` = lognormal(log(1000), 0.2),
        `2` = lognormal(log(1500), 0.3),
        `3` = lognormal(log(800),  0.25)
      ))
    )
  )
)
```

The prior contribution appears in slot 20 of the joint NLL:

```{r}
fit$quantities$jnll_comp["Linkage-table priors", ]
```

You can inspect the linked parameter values via the linkage table,
the estimated slope coefficients, and the base-parameter array:

```{r}
fit$data_list$linkage_table[fit$data_list$linkage_table$param == "K", ]
fit$estimated_params$beta_linkage          # (Intercept) rows are 0; slopes are estimated
fit$estimated_params$log_growth_pars       # base K / L1 / Linf / m (on log scale internally)
```

The same approach works for **`M1`** when you want a prior on
natural mortality itself:

```{r}
data("GOApollock") # Using GOA pollock as an example

# One-species model, lognormal(log(0.34), 0.01) prior on M1
# applied via the linkage intercept (equivalent to a Normal on log_M1).
# The structural M1_model is kept at "sex_age_invariant" (= 1).
M1_spec <- build_M1(
  M1_model = "sex_age_invariant",
  linkages = list(
    M1 = linkage_spec(
      formula = ~ 1,
      by      = ~ species,
      init    = list(`(Intercept)` = 0.34),
      # Prior is on the natural-scale for M1.
      # `lognormal(log(0.34), 0.01)` evaluates Normal on log(M1).
      # Use `normal(0.34, 0.01)` for a Normal directly on natural M1.
      priors  = list(`(Intercept)` = lognormal(log(0.34), 0.01))
    )
  )
)

# Fit the model - no manual mapping needed
fit <- fit_mod(
  data_list = GOApollock,
  M1Fun     = M1_spec,
  estimateMode = 0,
  fit_control = fit_control(phase = TRUE)
)
```

The linkage prior is applied verbatim, with no lognormal bias-correction term.

## Species- and (species, sex)-specific priors

Each entry under `priors = list(...)` may be:

* a single prior -- applies to every species/sex row that uses
  that design column,
* a **named list of priors keyed by species id** -- one prior per
  specified species, or
* a **named list of priors keyed by species id and sex** -- one prior per
 specified species and sex, or
* a **two-level nested list keyed first by specified species, then by specified sex
  id**.

```{r}
# Species-only keying
linkage_spec(
  formula = ~ temp,
  by      = ~ species,
  priors  = list(
    temp = list(
      `1` = normal(0, 0.1),    # tight prior for species 1
      #`2` = normal(0, 0.5),   # No prior species 2
      `3` = normal(0, 1.0)     # weakest for species 3
      # any species not listed gets no prior
    )
  )
)
```

For sexually dimorphic stocks, the same `temp` column can take
different priors per (species, sex). At each level the value can be
either an `Rceattle_prior` (applies to all sub-strata) or a list
that drills down further:

```{r}
linkage_spec(
  formula = ~ temp,
  by      = ~ species + sex,
  priors  = list(
    temp = list(
      # Species 1: different tightness per sex.
      `1` = list(`1` = normal(0, 0.1),    # sp 1, sex 1 (e.g. females)
                 `2` = normal(0, 0.2)),   # sp 1, sex 2 (e.g. males)
      # Species 2: scalar applies to every sex.
      `2` = normal(0, 0.5)
      # Any (species, sex) combination not listed gets no prior.
    )
  )
)
```

The same syntax works for any prior family. This is the
recommended pattern when a covariate has different prior strength
across stocks (e.g. when one species has prior literature on its
temp sensitivity but another doesn't), or when within a species
the sexes are expected to respond differently. Built
programmatically, it's just:

```{r}
sd_by_species <- c(`1` = 0.1, `2` = 0.5, `3` = 1.0)
priors_per_sp <- lapply(sd_by_species, function(s) prior_normal(0, s))

linkage_spec(
  formula = ~ temp,
  by      = ~ species,
  priors  = list(temp = priors_per_sp)
)
```


## Per-species formulas

If species have qualitatively different env relationships -- e.g.
species 1 responds only to temperature, species 2 also to PDO --
register *multiple specs against the same parameter* and use the
`species` argument on each to scope it to a subset of stocks:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    K = list(
      linkage_spec(
        formula = ~ temp,
        by      = ~ species,
        species = 1L,                                    # only sp 1
        priors  = list(temp = normal(0, 0.5))
      ),
      linkage_spec(
        formula = ~ temp + PDO,
        by      = ~ species,
        species = 2L,                                    # only sp 2
        priors  = list(temp = normal(0, 1),
                       PDO  = normal(0, 1))
      )
    )
  )
)
```


Linkages can accepts both forms under each parameter key:

```r
linkages = list(K = single_spec)             # one formula for all sp
linkages = list(K = list(spec_a, spec_b))    # different formulas per sp
```

Inspecting the result, `fit$linkages$K` is either an
`Rceattle_linkage_spec` or a `list` of them, 
and `fit$data_list$linkage_table` shows the
materialized rows with `species` and `X_col`.

This pattern composes with species-specific *priors* -- each spec
carries its own priors, and within a spec the priors can themselves
be species-keyed if you want even finer control:

```{r}
linkage_spec(
  formula = ~ temp,
  by      = ~ species + sex,
  species = c(1L, 2L),
  priors  = list(
    temp = list(
      `1` = normal(0, 0.3),
      `2` = normal(0, 0.7)
    )
  )
)
```

## Per-sex formulas

The `sex` argument on `linkage_spec()` mirrors `species`: it scopes a
spec to a subset of sex ids. Combined with the multi-spec list form,
this lets you register *separate specs per sex* against the same
parameter -- typical use cases are sex-dimorphic M priors or distinct
env responses for females vs. males:

```{r}
# Female and male M1 each get their own intercept-only prior.
M1_spec <- build_M1(
  M1_model = "sex_specific",
  linkages = list(
    M1 = list(
      linkage_spec(
        formula = ~ 1,
        by      = ~ species + sex,
        sex     = "Females",                              # or sex = 1L
        init    = list(`(Intercept)` = 0.12),
        priors  = list(`(Intercept)` = lognormal(log(0.12), 0.0001))
      ),
      linkage_spec(
        formula = ~ 1,
        by      = ~ species + sex,
        sex     = "Males",                                # or sex = 2L
        init    = list(`(Intercept)` = 0.20),
        priors  = list(`(Intercept)` = lognormal(log(0.20), 0.0001))
      )
    )
  )
)
```

The `sex` filter is a no-op when `by` does not include `sex`. If you
want different *formulas* per sex (rather than just different priors
or inits), use the multi-spec form -- one spec with `sex = 1L` and
its own RHS, another with `sex = 2L` and a different RHS. 

If you only need different priors per (species, sex) for a single
shared formula, the nested-priors form (see *Species- and (species,
sex)-specific priors* above) is simpler -- one spec, one set of
design columns, prior values keyed by `species` then `sex`. Reach for
the per-sex multi-spec form when you need different RHS terms,
different `init`/`bounds`, or different `est_phase` between sexes.

## Polynomial and basis-expansion formulas

The formula passed to `linkage_spec()` is fed straight to
[`stats::model.matrix()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/model.matrix.html),
so any expression `model.matrix()` understands works -- polynomial
bases, splines, interactions, and arithmetic transformations.

**Polynomials.** Use `poly()` for orthogonal polynomial bases (the
default; numerically well-conditioned), or `poly(..., raw = TRUE)`
for raw monomials. Each basis order shows up as a separate design
column with its own coefficient:

```{r}
# 4th-order orthogonal polynomial in temp, by species.
linkage_spec(
  formula = ~ poly(temp, 4),
  by      = ~ species
)
# design columns produced:
#   (Intercept), poly(temp, 4)1, poly(temp, 4)2,
#   poly(temp, 4)3, poly(temp, 4)4
```

`poly()` orthogonalises the basis against the rows of `env_data`,
so the columns are zero-mean over the model years and the linear
slope, quadratic curvature, etc. are estimated independently --
generally the right choice if you don't have a specific scientific
reason to prefer raw monomials.

**Arbitrary arithmetic.** Use `I()` to insert a literal expression
(otherwise `temp^2` is interpreted as a formula crossing operator
and reduces to `temp`):

```{r}
# Linear + raw quadratic in temp.
linkage_spec(
  formula = ~ temp + I(temp^2),
  by      = ~ species,
  priors  = list(`I(temp^2)` = normal(0, 0.1))   # tight prior on curvature
)
```

The design column name is the deparsed call (`I(temp^2)`), so when
you key `init`, `bounds`, or `priors` against it you have to use
the same string -- backticks let you write it as a name.

**Splines.** Natural cubic splines via `splines::ns()` work
out-of-the-box and are often a better choice than high-order
polynomials when you want flexibility without strong tail behaviour:

```{r}
linkage_spec(
  formula = ~ splines::ns(temp, df = 4),
  by      = ~ species
)
```

**Interactions.** `temp:PDO` adds a single product column;
`temp * PDO` is shorthand for `temp + PDO + temp:PDO`:

```{r}
linkage_spec(
  formula = ~ temp * PDO,
  by      = ~ species,
  priors  = list(`temp:PDO` = normal(0, 0.5))
)
```

**Mixing forms across species.** Combined with the per-species
formula machinery, you can give one species a polynomial response
and another a linear one without duplicating columns:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    K = list(
      linkage_spec(formula = ~ poly(temp, 3),
                   by = ~ species, species = 1L),
      linkage_spec(formula = ~ temp,
                   by = ~ species, species = 2L)
    )
  )
)
```


A practical caution: same-named columns must be numerically
identical across specs that use them. With deterministic
formulas evaluated against the same `env_data` this is automatic;
it can only fail if you've built specs against different
`env_data` frames.

## How the pieces fit together

When you fit, `fit_mod()` combines the linkages from every `build_*()`
call into a single linkage table and a shared design matrix (both stored
on `data_list`), then applies per-process offsets on top of the base
parameters each year. With no linkages the offsets are zero, so a model
without linkages fits exactly as before. Any priors and bounds you set on
a linkage flow through the same table and add to the joint NLL.

## Natural mortality

`build_M1()` supports the same `linkages` argument. The valid
parameter key is `M1`; the offset enters additively (on the log
scale) inside the `M1_at_age` compute, so a coefficient of `0.3` on
a normalised temperature index multiplies `M1` by `exp(0.3 *
temp[yr])` each year.

**Why `M1` and not `M`?** CEATTLE decomposes total natural mortality
as `M = M1 + M2`. Only `M1` (residual / non-predation) is a
parameter; `M2` (predation) is derived from predator abundance,
suitability, and ration each iteration. Environmental effects on
predation therefore propagate via upstream parameters (e.g. a temp
effect on predator `K`, on recruitment, or on bioenergetics ration
inputs).

```{r}
yrs <- GOApollock$styr:GOApollock$endyr
GOApollock$env_data <- data.frame(
  Year = yrs,
  temp = rnorm(length(yrs), 0, 0.1)
)

M1_spec <- build_M1(
  M1_model = "sex_age_invariant",            # or M1_model = 1 (back-compat)
  linkages = list(
    M1 = linkage_spec(
      formula = ~ temp,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.1))
    )
  )
)

fit <- fit_mod(
  data_list   = GOApollock,
  M1Fun       = M1_spec,
  estimateMode = 0,
  msmMode     = 0
)
```

The structural M choices (`M1_model`) and the random-effects
choices (`M1_re`) accept either integer codes or readable strings
interchangeably:

```{r}
build_M1(M1_model = "sex_age_specific", M1_re = "ar1_age")   # = 3, 4
build_M1(M1_model = 3,                  M1_re = 4)            # equivalent
```

`linkages` on M is age-aware: by default a row's `age_bin` is `NA`
and the offset broadcasts across every age. To pin an env effect to
a specific age slice, set `by = ~ species + age_bin` and supply
`strata$age_bin` when calling materialize directly (`fit_mod()`
provides this automatically). For most operational uses the default
broadcast is what you want.

The new linkage path coexists with the legacy `M1_indices = c(...)`
column-index argument for `M1_model = 4` / `5`: both add to
`ln_M1` on the log scale and either may be used. New code should
prefer `linkages` -- it composes naturally with priors, bounds, and
per-species formulas.

Inspecting the result:

```{r}
fit$data_list$linkage_table         # rows have process == "M"
fit$quantities$M_linkage_offset     # [nspp, max_sex, max_age, nyrs]
fit$quantities$jnll_comp["Linkage-table priors", ]
```

Growth and M linkages **share the same table** -- one fit can carry
linkage rows for both processes, and the underlying coefficient
vector `beta_linkage` indexes both.

## Recruitment

`build_srr()` accepts the same `linkages` argument with three
allowed parameter keys:

* **`R0`** -- meaningful for *any* `srr_fun`; the right target if
  recruitment should respond to env without fitting an SRR.
* **`alpha`** (productivity) and **`beta`** (density-dependence) --
  only do work when `srr_fun %in% c(2, 3, 4, 5)` (Beverton-Holt or
  Ricker).

```{r}
# Mean-recruitment with a temperature linkage on R0.
rec_spec <- build_srr(
  srr_fun  = "mean",                       # = 0
  linkages = list(
    R0 = linkage_spec(
      formula = ~ temp,
      priors  = list(temp = normal(0, 0.5))
    )
  )
)

# Beverton-Holt with env effects on alpha (and species-keyed
# priors so each stock has its own sensitivity).
rec_spec_bh <- build_srr(
  srr_fun  = "BevertonHolt",               # = 2
  linkages = list(
    alpha = linkage_spec(
      formula = ~ temp,
      priors  = list(
        temp = list(`1` = normal(0, 0.3),
                    `2` = normal(0, 0.7))
      )
    )
  )
)

fit <- fit_mod(
  data_list   = data_list,
  recFun      = rec_spec,
  estimateMode = 0,
  msmMode     = 0
)
```

The structural choice (`srr_fun`) accepts either integer codes or
readable strings interchangeably:

```{r}
build_srr(srr_fun = "mean")           # = 0
build_srr(srr_fun = "BevertonHolt")   # = 2
build_srr(srr_fun = "Ricker")         # = 4
```

Inspecting the result:

```{r}
fit$data_list$linkage_table             # rows have process == "recruitment"
fit$quantities$recruitment_linkage_offset  # [nspp, n_rec_params, nyrs]
                                           # n_rec_params = 3 (R0, alpha, beta)
```

Practical note: a linkage row's `age_bin` is ignored for
recruitment (recruitment is at age 0 by definition). The
`sex_ratio` data input partitions recruits across sexes downstream.

## Catchability

Survey/index catchability is indexed by **fleet**, not by
species/sex/age, so its linkages use `by = ~ fleet`. `build_catchability()`
takes the same `linkages` argument with a single parameter key, `q`:

```{r}
fit <- fit_mod(
  data_list = data_list,
  qFun      = build_catchability(
    linkages = list(
      q = linkage_spec(~ temp, by = ~ fleet)   # one coefficient per fleet
    )
  ),
  estimateMode = 0,
  msmMode      = 0
)

# One coefficient set per fleet; the offset lands here:
fit$quantities$q_linkage_offset      # [n_flt, nyrs]
```

Restrict a spec to particular fleets with the `fleet` argument of
`linkage_spec()`, and attach priors the same way as any other process:

```{r}
build_catchability(linkages = list(
  q = linkage_spec(~ temp, by = ~ fleet, fleet = c(1, 3),
                   priors = list(temp = normal(0, 1)))))
```

This **supersedes** the old `Catchability = "Environmental"` /
`Time_varying_q` column pair, which named covariates by their position in
a comma-separated list of `env_data` column indices. Existing models with
`Catchability = "Environmental"` still fit -- they keep their own code path
and exact numerics -- but emit a deprecation warning pointing at
`build_catchability()`. As a linkage the effect is written as a formula,
carries priors/bounds/phase, and names the fleet explicitly.

An intercept-bearing formula (`~ temp`) leaves the base `index_log_q`
estimable to carry the level and fixes the intercept row at 0, exactly as
the growth/M/recruitment intercepts re-target their base parameters.

### A prior on the survey catchability

Often you know something about a survey's catchability from outside the model:
an acoustic or diver survey believed to see close to the whole stock (`q ≈ 1`),
or a `q` prior borrowed from a catchability meta-analysis or a companion
assessment. Put that belief on the fleet's catchability with a `~ 1` formula
and a prior on the intercept:

```{r}
build_catchability(linkages = list(
  # Survey q for fleet 2 believed near 1, with moderate uncertainty.
  q = linkage_spec(~ 1, by = ~ fleet, fleet = 2,
                   priors = list(`(Intercept)` = lognormal(log(1), 0.2)))))
```

The prior is on `q` on its natural scale, so a `lognormal(log(1), 0.2)` says
"`q` is near 1, and I'd be mildly surprised by a factor of ~1.5 either way."
`~ 1` alone gives an estimated `q` with a prior on its level and no
environmental effect; write `~ 1 + temp` to keep the level prior *and* link
`q` to a covariate.

## Selectivity

Selectivity is also fleet-indexed (`by = ~ fleet`) and, unlike the other
processes, has several parameters. `build_selectivity()` accepts these
keys, which name the underlying slots shared across the parametric forms:

| key | logistic family | DoubleNormal |
|---|---|---|
| `slp_asc`, `slp_desc` | ascending / descending slope | `sigma_asc`, `sigma_desc` |
| `inf_asc`, `inf_desc` | ascending / descending inflection | `peak`, `right_floor` |

The DoubleNormal aliases (`sigma_asc`, `peak`, ...) map to the same slots,
so you name the quantity your fleet's form actually has.

```{r}
fit <- fit_mod(
  data_list = data_list,
  selFun    = build_selectivity(
    linkages = list(
      inf_asc = linkage_spec(~ cold_pool, by = ~ fleet)
    )
  ),
  estimateMode = 0,
  msmMode      = 0
)
```

As with every process, `link = "log"` makes the effect multiplicative on
the natural parameter and `link = "identity"` makes it additive, so you can
link the log-space slope and the natural-space inflection independently.

**Supported forms:** the parametric selectivities -- `Logistic`,
`DoubleLogistic`, `DescendingLogistic`, `DoubleNormal`, and `LogisticPM`.
A linkage on any other form errors at fit time naming the fleet, rather than
being estimated to no effect. The non-parametric forms (`coff`) are
excluded on principle: they mean-centre their coefficients every year, so a
per-year offset applied across all bins cancels exactly -- a per-bin
covariate would be needed, which the formula grammar does not express.
`Time_varying_sel` process-error deviates are a separate mechanism and
compose additively with a selectivity linkage.

## Data weighting (Dirichlet-multinomial)

The Dirichlet-multinomial (DM) likelihood (Thorson et al. 2017) is a
self-weighting alternative to Francis or McAllister–Ianelli tuning: instead of
iterating an external weight, it estimates each composition dataset's
**effective sample size** as part of the fit. That works well when a fleet has
many comp years, but the weight can be poorly identified when a fleet has only
a handful -- the fit may push the effective sample size to an implausible
extreme. A prior on the DM weight keeps it in a believable range without
reverting to hand-tuning:

```{r}
# Use the DM likelihood on the fleets you want it to self-weight ...
data_list$fleet_control$Comp_loglike[c(1, 2)] <- "DirichletMultinomial"

# ... and (optionally) constrain its weight with a prior.
fit <- fit_mod(
  data_list = data_list,
  compFun   = build_composition(
    linkages = list(
      theta_comp = linkage_spec(
        ~ 1, by = ~ fleet, fleet = c(1, 2),
        # Gamma(2, 0.5): mean 4, mildly informative. Keeps the weight off 0
        # (comps ignored) and off very large (comps fit as if unweighted).
        priors = list(`(Intercept)` = gamma(2, 0.5)))
    )
  ),
  estimateMode = 0,
  msmMode      = 0
)
```

Three keys, one per kind of composition data:

| key | applies to | DM likelihood set on |
|---|---|---|
| `theta_comp` | age / length composition | `Comp_loglike` |
| `theta_caal` | conditional age-at-length | `CAAL_loglike` |
| `theta_diet` | predator diet composition | `Diet_loglike` |

The prior is on the weight itself (a larger weight = a larger effective sample
size = the composition data trusted more). It only makes sense where the DM is
actually estimating that weight, so a `theta_*` linkage on a fleet or predator
whose likelihood is **not** `DirichletMultinomial` errors up front rather than
sitting inert. And because the weight is a single number per fleet, not a
time series, this is a **prior only**: a covariate (`~ temp`), an `init`, or an
`est_phase` on a composition spec is rejected -- start or fix the weight
through the model's usual `inits` / `map` if you need to.

## A multi-process example

The same pattern composes when more than one parameter is linked:

```{r}
growth_spec <- build_growth(
  fun = "Richards",
  linkages = list(
    K    = linkage_spec(
      formula = ~ temp,
      by      = ~ species + sex,
      priors  = list(temp = normal(0, 1))
    ),
    Linf = linkage_spec(
      formula = ~ temp + PDO,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.5),
                     PDO  = normal(0, 0.5))
    ),
    m    = linkage_spec(
      formula = ~ 1,             # intercept only -- no env effect
      by      = ~ species
    )
  )
)
```

The pooler will:

* compute one shared `X = model.matrix(~ temp + PDO, env_data)` of
  three columns,
* emit 4 (sp × sex) + 3 (Linf cols) × 3 (sp) + 3 (m intercepts) = 16
  rows in the linkage table,
* match same-named columns (`(Intercept)`, `temp`) across specs so
  there's no duplication in `X`.

Per-process logic (vB has no `m`; only Richards does) is enforced at
build time:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(m = linkage_spec(~ temp))
)
#> Error: linkages$m is only valid with fun = 'Richards'; ...
```

Growth, M, and recruitment can all be linked in the same fit --
each `build_*()` output carries its own linkages list and
`fit_mod()` pools them against a single shared design matrix:

```{r}
fit <- fit_mod(
  data_list   = data_list,
  growthFun   = build_growth(
    fun      = "vonBertalanffy",
    linkages = list(
      K = linkage_spec(formula = ~ temp)
    )
  ),
  M1Fun       = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(formula = ~ temp)
    )
  ),
  recFun      = build_srr(
    srr_fun  = "mean",
    linkages = list(
      R0 = linkage_spec(formula = ~ temp)
    )
  ),
  estimateMode = 0,
  msmMode     = 0
)

# Linkage table contains rows for all three processes:
table(fit$data_list$linkage_table$process)
#>      growth           M  recruitment
#>           4           4            4
```

## Inspecting and debugging

A few diagnostics worth knowing:

```{r}
# The materialised table -- one row per estimated coefficient.
fit$data_list$linkage_table

# The shared design matrix -- one column per unique RHS term.
fit$data_list$linkage_X

# Per-(species, sex, growth_param, year) offsets actually applied to
# growth_parameters this iteration. Useful for verifying the linkage
# fired as you intended.
str(fit$quantities$growth_linkage_offset)
str(fit$quantities$M_linkage_offset)
str(fit$quantities$recruitment_linkage_offset)

# Per-row coefficient values in the same order as `linkage_table`.
fit$estimated_params$beta_linkage

# The prior NLL contribution from slot 19 of jnll_comp.
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## Behind the scenes: automatic base-parameter handling

For three species × two sexes × `~ temp` (intercept + slope) the
pooled table has 12 rows -- an `(Intercept)` and a `temp` slope per
(species, sex). Priors and inits supplied for `(Intercept)` flow to
the *base parameter* (`K`, `M1`, `R0`, ...); the slope rows hold the
year-by-year offset that combines with the base. The exact behavior
depends on whether the formula carries an intercept:

| Formula                      | Base parameter                            | Slope rows |
|------------------------------|-------------------------------------------|------------|
| `~ 1` (intercept only)       | estimable; init from spec or default      |n/a |
| `~ temp` (intercept + slope) | estimable; init from spec or default      | estimated |
| `~ 0 + temp` (slope only)    | mapped NA, value from spec or default | estimated |

**Intercept-bearing formulas (`~ 1`, `~ temp`)** -- the base parameter
remains estimable; for example `R0` for
species `s` lives on `rec_pars[s, 1]` exactly as it does without any
linkages. The `(Intercept)` row of the table exists for bookkeeping
and as a hook for `init`/`prior` (see below) but its `beta_linkage`
slot is held at `0` and not estimated. The recruitment / mortality /
growth modules read `base + offset(yr)` each year -- with the
intercept fixed at zero, the offset is exactly the slope contribution.

**Slope-only formulas (`~ 0 + temp`)** -- there is no `(Intercept)`
row, so the base parameter is mapped out (NA) and stays at the
default initial value, or the value supplied by the `inits` argument in `fit_mod`, or 
the value supplied by `init = list("(Intercept)"= ...)` on the spec.

### `init` and the base parameter

`init = list(`(Intercept)` = X)` on the linkage spec **flows to the
base parameter** rather than to the linkage row. The linkage row's
`beta_linkage` slot stays at `0`; the base parameter starts at `X`.
This is the recommended way to set a sensible starting value for
intercept-bearing linkages -- it composes naturally with phasing
(the base parameter goes through the existing per-process phase
logic) and keeps the optimiser grounded.

```{r}
# Start M1 at 0.06 without estimating an extra intercept beta.
build_M1(
  M1_model = "sex_age_invariant",
  linkages = list(
    M1 = linkage_spec(
      formula = ~ 1,
      init    = list(`(Intercept)` = 0.06)
    )
  )
)
```

### Stratification

Stratification (`by`) determines *which* base-parameter slots are
affected: `by = ~ species` produces one `(Intercept)` row per species;
`by = NULL` (a fully shared linkage) produces a single shared row;
and `by = ~ species + sex` produces one per (species, sex). Each
`(Intercept)` row that the user supplies an `init` for pushes that
init into the matching base-parameter slot.

This rule applies uniformly to every linkage target (`K`, `L1`,
`Linf`, `m`, `M1`, `R0`, `alpha`, `beta`).

### Single-sex models and `by = ~ ... + sex`

If your model is single-sex (`nsex = 1` for every species in scope),
asking for `by = ~ species + sex` produces a warning: there is only
one sex level available, so the sex stratum collapses to a single
shared level. The result is equivalent to `by = ~ species`. Models
with mixed `nsex` (some single-sex, some two-sex species) emit one
sex level per species automatically -- single-sex species get one row
per parameter, two-sex species get two.

## Random effects

A linkage formula may carry a **random effect**, so a parameter varies
year-to-year (or over any grouping) as a set of deviations damped by an
estimated density rather than fixed covariate slopes:

* `~ (1 | Year)` -- **IID** deviations, `N(0, sigma)`;
* `rw(1 | Year)` -- a **random walk**, `N(0, sigma)` on first differences
  (first deviate pinned so the base parameter carries the level);
* `ar1(1 | Year)` -- a stationary **AR1**, `SCALE(AR1(rho), sigma)` with
  `sigma` the marginal SD and `rho` the correlation.

Note that `sigma` means different things across structures. For `rw()` it is
the **innovation** (step) SD -- the SD of each `year-to-year` change -- because
a random walk is non-stationary and has no overall spread. For `ar1()` it is
the **marginal** (stationary) SD -- the overall spread of the deviations
(`sigma_marginal = sigma_innovation / sqrt(1 - rho^2)`) -- following the
`glmmTMB` convention. The two agree at `rho = 0` (where `ar1` reduces to `IID`),
but if you switch a group from `rw()` to `ar1()` the estimated `sigma` is not
directly comparable across the two.

Each grouping level gets one deviation; each group estimates its own SD. The
SD and (for `ar1`) the correlation route through the same `init` / `priors`
arguments as every other parameter: `init = list(sigma = 0.1)` fixes the SD
(reproducing the legacy `Time_varying_*_sd_prior` input), `init = list(rho =
0.7)` fixes the correlation, and `priors = list(sigma = lognormal(...))` /
`priors = list(rho = normal(...))` place priors. `rw()` / `ar1()` need a
numeric grouping variable (a real elapsed-time lag).

An `ar1(1 | Year)` term can additionally be a **state-space covariate** (the
Rogers et al. 2024 QAR1 catchability model): with `observe = "<env column>",
obs_sd = <value>` the AR1 latent is observed as an `env_data` column with a
fixed measurement SD and enters the linked parameter through an estimated
effect size. The effect size and the AR1 SD are separately identified only when
`obs_sd` is informative (small relative to the latent's SD); an uninformatively
large `obs_sd` lets the observation stop pinning the latent, so use the
measured series' actual measurement error.

`rw()` / `ar1()` couple *consecutive* deviations with a unit lag, so their time
points must be **equally spaced** -- a gap (a missing year) would be silently
treated as a single step. A grid with gaps errors rather than mis-specifying
the process.

`env_data` is applied to model years **by row position**: row 1 is the model
start year, and years beyond the last row get a zero offset (so `env_data` may
stop short of the projection horizon). When `env_data` carries a `Year` column
it must start at `styr` and be contiguous; a misaligned or gappy `Year` errors
rather than applying a covariate/deviate to the wrong year.

## On the roadmap

* **Legacy translation** -- `Time_varying_q` / `Time_varying_sel` / `M1_re`
  configurations are not yet auto-translated onto the equivalent random-effect
  grammar; both paths coexist. The `Time_varying_*_sd_prior` column is an input
  SD (reproduced by `init = list(sigma = )`), not a prior, despite its name.
* **Age-dependent linkages on processes other than M** -- the
  `age_bin` stratum is consumed by the M accumulator today (a
  row's `age_bin == NA` broadcasts across ages; specific values
  pin the offset to that age slice). The schema is reserved for
  other processes that may eventually vary by age.
* **Non-parametric selectivity** -- a `coff` linkage needs a per-bin
  covariate, since a per-year scalar cancels under the annual
  mean-centring; the formula grammar does not yet express that.
* **Legacy retirement** -- `srr_indices`, `M1_indices`,
  `srr_fun %in% c(1, 3, 5)`, and `M1_model %in% c(4, 5)` are
  deprecated; see the migration table in `NEWS.md`.

## See also

* `?linkage_spec`, `?build_growth`, `?build_M1`, `?build_srr`,
  `?build_catchability`, `?build_selectivity`, `?build_composition`,
  `?prior_normal`
* `vignette("model-parameterizations", package = "Rceattle")` for the
  full set of process options.
