---
title: "Environmental linkages: a formula-driven, long-format API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Environmental linkages: a formula-driven, long-format API}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Why this exists

Environmental drivers (temperature, prey density, climate indices) enter
stock assessments through several distinct process modules: recruitment,
natural mortality, growth, catchability, selectivity. Historically each
of these modules wires its own column-index argument
(`srr_indices`, `M1_indices`, `growth_indices`, ...) into a wide
parameter array (`beta_rec_pars[nspp, n_env]`,
`M1_beta[nspp, nsex, n_env]`, ...). That pattern doesn't scale: every
new process invents its own dimensionality, the parameter arrays end up
mostly empty when only a handful of links are estimated, and adding a
new combination (e.g. *log_K varies by sex AND temperature*) requires
new code in three or four places.

`Rceattle` ships an alternative: a **long-format linkage table** where
each row of the table corresponds to exactly one estimated coefficient.
Users describe the linkages they want with `linkage_spec()` (a
formula-driven helper) and pass the result into the relevant
`build_*()` function. Internally, `fit_mod()` pools every spec into a
single `data.frame` with one row per `beta`, paired with a shared
design matrix `X`. The TMB template iterates the table once,
accumulates per-process offsets, and adds them to the underlying linear
predictors.

This vignette walks through the user-facing API, the underlying data
structures, and the prior / bound options. Today the long-format path
is fully wired for **growth** (von Bertalanffy and Richards). The same
machinery is being prepared for natural mortality and recruitment.

## A first example: temperature on `K`

Suppose we want sex- and species-specific von Bertalanffy growth, with
the von Bertalanffy `K` parameter responding to a temperature index by
species:

```{r}
library(Rceattle)
data(BS2017SS)

# 1. Attach an env_data table to the data list. Must include `Year`
#    plus whatever covariates show up in your formulas. Here `temp` is
#    a placeholder; in real use it would be a standardised SST series,
#    PDO index, etc.
yrs <- BS2017SS$styr:BS2017SS$endyr
BS2017SS$env_data <- data.frame(
  Year = yrs,
  temp = scale(seq_along(yrs))[, 1]
)

# 2. Build a growth specification with one linkage per parameter that
#    you want to vary. linkage_spec() captures (formula, by, link,
#    init, bounds, priors) for one process parameter.
growth_spec <- build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = linkage_spec(
      formula = ~ temp,             # 2 columns: (Intercept) and `temp`
      by      = ~ species + sex,    # one beta per (species, sex, column)
      link    = "log",              # K lives on the log scale already
      priors  = list(temp = normal(0, 1))
    )
  )
)

# 3. Fit as usual.
fit <- fit_mod(
  data_list   = BS2017SS,
  growthFun   = growth_spec,
  estimateMode = 0,
  msmMode     = 0,
  fit_control = fit_control(verbose = 0)
)
```

The pooled linkage table is stored on the fitted model:

```{r}
fit$data_list$linkage_table
#>      process    param species sex age_bin X_col link  init prior_family ...
#> 1     growth    log_K       1   1      NA     1 log   0.0         none ...
#> 2     growth    log_K       1   2      NA     1 log   0.0         none ...
#> 3     growth    log_K       2   1      NA     1 log   0.0         none ...
#> 4     growth    log_K       2   2      NA     1 log   0.0         none ...
#> 5     growth    log_K       1   1      NA     2 log   0.0       normal ...
#> 6     growth    log_K       1   2      NA     2 log   0.0       normal ...
#> 7     growth    log_K       2   1      NA     2 log   0.0       normal ...
#> 8     growth    log_K       2   2      NA     2 log   0.0       normal ...
```

For three species × two sexes × `~ temp` (intercept + slope) you get
12 rows: an intercept beta and a temp-slope beta per (species, sex)
combination. The intercept rows aren't a duplicate of `ln_growth_pars`
-- they are just zero-valued columns of the global design matrix and
typically stay at their initial value (you can use the linkage table
solely for slopes if you prefer).

The shared design matrix is stored alongside:

```{r}
head(fit$data_list$linkage_X)
#>    (Intercept)        temp
#> 1            1  -1.5933919
#> 2            1  -1.4658285
#> 3            1  -1.3382651
#> ...
```

## The `linkage_spec()` API

```r
linkage_spec(
  formula,
  param     = NULL,
  by        = NULL,
  link      = c("identity", "log", "logit"),
  init      = NULL,
  bounds    = NULL,
  priors    = NULL,
  re_group  = NA_character_,
  est_phase = 1L
)
```

* **`formula`** -- a one-sided R formula whose RHS describes the
  linear predictor for the target parameter. Anything `model.matrix()`
  understands works: `~ 1`, `~ temp`, `~ temp + PDO`, `~ temp:PDO`,
  `~ poly(temp, 2)`, factor predictors, etc.
* **`param`** -- the target parameter name (e.g. `"log_K"`). When the
  spec is registered inside a `build_*()` linkages list keyed by
  parameter name, `param` is filled in automatically from the list
  key.
* **`by`** -- a one-sided formula naming the stratifying factors that
  should each get their own coefficients. Allowed names are
  `species`, `sex`, and `age_bin`. `by = ~species` produces one
  coefficient set per species; `by = ~species + sex` produces one
  per (species, sex) cell. `NULL` means a single coefficient applies
  to every species/sex.
* **`link`** -- the link function applied to the predictor when the
  offset is added back to the underlying parameter. `"identity"` and
  `"log"` are equivalent for parameters already on a log scale (which
  is currently true for all growth parameters). `"logit"` is reserved
  for future use (e.g. priors on steepness in `(0, 1)`).
* **`init`**, **`bounds`** -- optional named lists keyed by
  design-matrix column name. `init = list(temp = 0.05)` sets the temp
  slope's initial value to 0.05; `bounds = list(temp = c(-2, 2))`
  constrains it during optimisation.
* **`priors`** -- a named list of prior objects keyed by design-matrix
  column. Inside `linkage_spec()` (and only there) the unprefixed
  short-names `normal()`, `lognormal()`, `gamma()`, and `beta()`
  resolve to the corresponding `prior_*` constructors via a private
  data mask -- see *Priors* below.
* **`re_group`** -- random-effects grouping label (currently a
  placeholder).
* **`est_phase`** -- estimation phase ordinal. `0` fixes the
  coefficient at its initial value; `1+` makes it estimable. Phased
  fitting through `fit_control(phase = TRUE)` honors the values you
  pick.

## Priors

Priors are first-class. Five families are available out of the box:

```r
prior_normal(mean, sd)
prior_lognormal(meanlog, sdlog)
prior_gamma(shape, rate)
prior_beta(shape1, shape2)
# Family "none" is implicit -- no prior contribution.
```

Inside `linkage_spec(priors = ...)` the same constructors are
available **without the `prior_` prefix** so user code stays close to
mathematical notation:

```{r}
linkage_spec(
  formula = ~ temp + PDO,
  by      = ~ species,
  priors  = list(
    temp = normal(0, 1),     # informative prior, ~95% in (-2, 2)
    PDO  = normal(0, 5)      # weak prior
  )
)
```

The unprefixed forms are **scoped** to the `priors = ...` argument and
do not mask `base::gamma()` or `base::beta()` at the package level.
Outside that argument you can still call the math functions directly:

```{r}
gamma(5)       # 24, the math gamma function
beta(2, 3)     # 1/12, the math beta function

linkage_spec(
  formula = ~ temp,
  priors  = list(temp = gamma(2, 3))   # interpreted as prior_gamma(shape=2, rate=3)
)
```

For programmatic prior assembly (e.g. species-specific priors built in
a loop) call the `prior_*` constructors directly:

```{r}
priors <- lapply(species_specific_sds, function(s) prior_normal(0, s))
names(priors) <- c("temp", "PDO", "salinity")
linkage_spec(formula = ~ temp + PDO + salinity, priors = priors)
```

Each prior's negative log-density contributes to slot 19 of the
joint NLL (`fit$quantities$jnll_comp[20, ]` in R, since R is
1-indexed):

```{r}
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## What happens internally

When you call `fit_mod()` with linkages:

1. **`build_growth(linkages = ...)`** returns a list whose
   `linkages` element is a named list of `Rceattle_linkage_spec`
   objects. Backwards-compat fields `growth_model`,
   `growth_re`, `growth_indices` are still populated with sensible
   defaults so the existing TMB dispatch (vB / Richards / empirical)
   keeps working.

2. **`fit_mod()`** stashes `growth_fun` and `growth_linkages` on
   `data_list`, runs `data_check()`, then calls `pool_linkages()`.

3. **`pool_linkages()`** materialises every spec against
   `data_list$env_data` and the species/sex/age strata, unions the
   per-spec design columns by name into a single shared design matrix
   `X`, and remaps each row's local `X_col` to the global column
   index. The resulting `Rceattle_linkage_table` and `linkage_X`
   matrix are stored on `data_list`.

4. **`encode_linkage_for_tmb()`** turns the table into the parallel
   integer/numeric vectors that the TMB template consumes
   (`linkage_process`, `linkage_param`, `linkage_species`,
   `linkage_sex`, `linkage_age_bin`, `linkage_X_col`,
   `linkage_link`, `linkage_prior_family`, `linkage_prior_p1`,
   `linkage_prior_p2`).

5. **The TMB template** iterates the linkage rows once, builds a
   per-(species, sex, growth_param, year) offset tensor via
   `rceattle_apply_linkages()` in `src/TMB/linkage.hpp`, and adds
   each cell to the corresponding entry of `growth_parameters` on
   the log scale. With no linkages the offset stays at zero so
   existing fits are identical-by-construction.

6. **Priors and bounds** flow through the same table:
   `build_bounds()` reads each row's `lower` / `upper` into
   `bounds$lower$ln_beta_linkage` / `bounds$upper$ln_beta_linkage`,
   and the cpp adds `-d<family>(beta, p1, p2, log = TRUE)` to slot
   19 of the joint NLL for every row whose `prior_family` is not
   `"none"`.

## A multi-process example

The same pattern composes when more than one parameter is linked:

```{r}
growth_spec <- build_growth(
  fun = "Richards",
  linkages = list(
    log_K    = linkage_spec(
      formula = ~ temp,
      by      = ~ species + sex,
      priors  = list(temp = normal(0, 1))
    ),
    log_Linf = linkage_spec(
      formula = ~ temp + PDO,
      by      = ~ species,
      priors  = list(temp = normal(0, 0.5),
                     PDO  = normal(0, 0.5))
    ),
    log_m    = linkage_spec(
      formula = ~ 1,             # intercept only -- no env effect
      by      = ~ species
    )
  )
)
```

The pooler will:

* compute one shared `X = model.matrix(~ temp + PDO, env_data)` of
  three columns,
* emit 4 (sp × sex) + 3 (Linf cols) × 3 (sp) + 3 (m intercepts) = 16
  rows in the linkage table,
* match same-named columns (`(Intercept)`, `temp`) across specs so
  there's no duplication in `X`.

Per-process logic (vB has no `m`; only Richards does) is enforced at
build time:

```{r}
build_growth(
  fun = "vonBertalanffy",
  linkages = list(log_m = linkage_spec(~ temp))
)
#> Error: linkages$log_m is only valid with fun = 'Richards'; ...
```

## Inspecting and debugging

A few diagnostics worth knowing:

```{r}
# The materialised table -- one row per estimated coefficient.
fit$data_list$linkage_table

# The shared design matrix -- one column per unique RHS term.
fit$data_list$linkage_X

# Per-(species, sex, growth_param, year) offsets actually applied to
# growth_parameters this iteration. Useful for verifying the linkage
# fired as you intended.
str(fit$quantities$growth_linkage_offset)

# Per-row coefficient values in the same order as `linkage_table`.
fit$estimated_params$ln_beta_linkage

# The prior NLL contribution from slot 19 of jnll_comp.
fit$quantities$jnll_comp["Linkage-table priors", ]
```

## What's not yet wired

* **Mortality and recruitment** -- the long-format pipeline is in
  place but the existing `build_M1(M1_indices = ...)` and
  `build_srr(srr_indices = ...)` paths are still the canonical way
  to express environmental linkages on `M` and `R`. Migration to
  `linkages = list(...)` is planned.
* **Random-effects grouping** -- the `re_group` column on the linkage
  table is stored end-to-end but the cpp does not yet introduce a
  hierarchical penalty. For now every row is a fixed effect.
* **Age-dependent linkages** -- the `age_bin` stratum is reserved
  in the table schema but not yet consumed by any process.

These are tracked alongside the rest of the post-release roadmap.

## Migrating from `growth_indices`

The legacy interface still works for environmental effects on
recruitment and mortality:

```{r}
# Legacy (M1 only, still supported)
build_M1(M1_model = 4, M1_indices = c(2, 3))
```

For growth, switch from `growth_indices` to `linkages`:

```{r}
# Old (no longer supported -- growth_indices was never plumbed to TMB):
# build_growth(growth_model = 1, growth_indices = c(2, 3))

# New
build_growth(
  fun = "vonBertalanffy",
  linkages = list(
    log_K = linkage_spec(formula = ~ env2 + env3, by = ~ species)
  )
)
```

The string and integer forms of `fun` are interchangeable
(`fun = 1` is shorthand for `fun = "vonBertalanffy"`) so existing
scripts using integer codes need no changes apart from substituting
`fun =` for `growth_model =`.

## See also

* `?linkage_spec`, `?build_growth`, `?prior_normal`
* `vignette("model-parameterizations", package = "Rceattle")` for the
  full set of process options.
