---
title: "2. HCRs and MSEs: an introduction"
author: "Grant Adams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{2. HCRs and MSEs: an introduction}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Fitting the operating model

A management strategy evaluation (MSE) runs a management strategy in closed loop: an operating model (OM) represents the "true" stock and generates data, an estimation model (EM) is fitted to that data and applies a harvest control rule (HCR), and the resulting catch feeds back into the OM. This example uses the northern rockfish assessment as the OM.

```{r, message = FALSE, warning = FALSE,echo = TRUE}
library(Rceattle)
data("NorthernRockfish2022")
nrdata <- NorthernRockfish2022
nrdata$fleet_control$Proj_F_proportion <- 1

om <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(
        formula = ~ 1,
        priors = list(
          "(Intercept)" = lognormal(log(0.06), 0.05)
        )
      )
    )
  ),
  fit_control = fit_control(phase = TRUE)
)
```

------------------------------------------------------------------------

## Fitting the estimation model (Tier 3 HCR)

The `HCR = build_hcr()` argument to `fit_mod()` sets the harvest control rule and its biological reference points (BRPs). Here we use the NPFMC Tier 3 rule: fishing at F40% (the max FABC proxy), an F35% overfishing limit (FOFL), and no fishing once spawning biomass drops below 20% of unfished. See `?build_hcr` for the full set of rules and reference points.

```{r, message = FALSE, echo = TRUE}
em <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0, # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(
        formula = ~ 1,
        priors = list(
          "(Intercept)" = lognormal(log(0.06), 0.05)
        )
      )
    )
  ),
  HCR = build_hcr(HCR = "NPFMC", # Tier3 HCR
                  Ftarget = 0.4, # F40%
                  Flimit = 0.35, # F35%
                  Plimit = 0.2,  # No fishing when SB<SB20
                  Alpha = 0.05),
  fit_control = fit_control(verbose = 1, phase = TRUE)
)
```

------------------------------------------------------------------------

## Running the MSE

`run_mse()` runs the closed loop: it projects the OM, samples data on the schedule you specify, re-fits the EM every `assessment_period` years, and applies the EM's HCR back to the OM. `nsim` sets the number of stochastic trajectories.

```{r, message = FALSE, echo = TRUE}
mse1 <- run_mse(
  om = om, em = em,
  nsim = 10,
  assessment_period = 2,
  # Fishery = fleet 1 (annual), BTS = fleet 2 (biannual)
  sampling_period = c(1, 2),
  seed = 666 # default; per-sim seed = seed + sim
)
```

### Reproducibility

`run_mse()` accepts two seed arguments, both with default `seed = 666`:

* `seed` controls the random draws *inside* each simulated trajectory
  (recruitment, observation error, composition resampling). Each
  simulation worker calls `set.seed(seed + sim)`, so simulation `1`
  uses seed `667`, simulation `2` uses `668`, and so on. Re-running
  the same `run_mse()` call with the same `seed` reproduces the
  trajectory bit-for-bit.
* `regenerate_seed` (defaults to `seed`) controls the optional
  `regenerate_past = TRUE` step that re-fits the EM to OM-simulated
  historical data. Set this separately if you want to vary
  past-data realizations while holding the projection draws fixed.

You do not need to call `set.seed()` before `run_mse()` — the `seed`
arguments fully control reproducibility. (You may still want
`set.seed()` for plotting code that draws random colours or jitter.)

For management-facing runs, record `packageVersion("Rceattle")` and
the `seed` alongside the saved `.rds` so a later reviewer can rerun
the exact same trajectories from the same package tag.

------------------------------------------------------------------------

## Evaluating the MSE with plots

Plotting the OM's spawning-biomass depletion across simulations shows the range of stock trajectories the strategy produces.

```{r, echo = TRUE}
plot_depletionSSB(lapply(mse1, function(x) x$OM))
```

```{r, echo = TRUE}
plot_depletionSSB(
  c(mse1$Sim_1$EM,           # EMs
    list(mse1$Sim_1$OM,      # OM
         mse1$Sim_1$OM_no_F  # OM w/ no fishing
    )),    
  line_col = c(
    paste0("grey", round(seq(80, 50, length.out = 15))),
    1, 3
  )
)
```

------------------------------------------------------------------------

## Summary statistics

`mse_summary()` reduces the trajectories to performance statistics for comparing strategies.

```{r, warning = FALSE, echo = TRUE}
mse_summ <- mse_summary(mse1)
knitr::kable(mse_summ, digits = 2)
```

------------------------------------------------------------------------

# Alternative scenarios

The runs below vary one element of the strategy at a time — the reference points, the survey frequency, the assessment interval, or the environment — so their performance statistics can be compared against `mse1`.

------------------------------------------------------------------------

## Different reference points

Here the EM uses a more aggressive rule (F30% target, F10% limit, no biomass cutoff) to contrast with the Tier 3 run above.

```{r, message = FALSE, echo = TRUE}
em2 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0, # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(
        formula = ~ 1,
        priors = list(
          "(Intercept)" = lognormal(log(0.06), 0.05)
        )
      )
    )
  ),
  HCR = build_hcr(HCR = "NPFMC", # Tier3 HCR
                  Ftarget = 0.30,# F30%
                  Flimit = 0.10, # F10%
                  Plimit = 0,    # No limit
                  Alpha = 0.1),
  fit_control = fit_control(verbose = 1, phase = TRUE)
)

mse2 <- run_mse(
  om = om, em = em2, 
  nsim = 10, 
  assessment_period = 2,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## Triennial bottom-trawl survey

`sampling_period = c(1, 3)` samples the fishery every year but the bottom-trawl survey (BTS) only every third year, testing the cost of a sparser survey.

```{r, message = FALSE, echo = TRUE, eval = FALSE}
mse3 <- run_mse(
  om = om, em = em, 
  nsim = 10, 
  assessment_period = 2,
  sampling_period = c(1, 3) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## 4-year assessment cycle and reduced survey coverage

Here the EM is re-fitted only every fourth year (`assessment_period = 4`) and `fut_sample = 0.5` halves the future sampling effort — a lower-cost monitoring scenario.

```{r, message = FALSE, echo = TRUE, eval = FALSE}
mse4 <- run_mse(
  om = om, em = em, 
  nsim = 10, 
  fut_sample = 0.5,
  assessment_period = 4,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## Climate-linked MSE

To test a strategy under environmental change, link a process in the OM to a covariate. Here recruitment (via `R0`) tracks a steadily rising environmental index, so the OM's productivity shifts over time while the EM keeps assuming stationary recruitment.

### Climate-linked OM

```{r, message = FALSE, echo = TRUE}

# Generate random env data
projyrs <- nrdata$styr:nrdata$projyr
projnyrs <- length(projyrs)

# Env data
nrdata$env_data <- merge(nrdata$env_data, data.frame(Year = projyrs, EnvIndex = seq(0,1, length.out = projnyrs)))

# Fit OM
om2 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    M1_model = "sex_age_invariant",
    linkages = list(
      M1 = linkage_spec(
        formula = ~ 1,
        priors = list(
          "(Intercept)" = lognormal(log(0.06), 0.05)
        )
      )
    )
  ),
  recFun = build_srr(
    srr_fun = "mean",    # Mean R with covariates
    proj_mean_rec = FALSE,
    linkages = list(
      R0 = linkage_spec(formula = ~ EnvIndex)
    )),
  fit_control = fit_control(phase = TRUE)
)

plot_biomass(om2, incl_proj = TRUE)
```

### Run MSE

```{r, message = FALSE, echo = TRUE, eval = FALSE}
mse5 <- run_mse(
  om = om2, em = em, 
  nsim = 10, 
  fut_sample = 0.5,
  assessment_period = 4,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

# Overall summary

Tagging each summary with an `MSE` label and stacking them puts the strategies side by side in one table.

```{r, echo = TRUE}
mse_summ$MSE = 1
mse_summ2 <- mse_summary(mse2); mse_summ2$MSE = 2

knitr::kable(do.call("rbind", list(mse_summ, mse_summ2)) |>
               dplyr::relocate(MSE), digits = 2)
```
