---
title: "2. HCRs and MSEs: an introduction"
author: "Grant Adams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{2. HCRs and MSEs: an introduction}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Fitting OMs

```{r, message = FALSE, warning = FALSE,echo = TRUE}
library(Rceattle)
data("NorthernRockfish2022")
nrdata <- NorthernRockfish2022
nrdata$fleet_control$proj_F_prop <- 1

om <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  fit_control = fit_control(phase = TRUE)
)
```

------------------------------------------------------------------------

## Fitting EMs (Tier 3 HCR)

We can use the `HCR = build_hcr()` argument in `fit_mod` to specify the HCR and BRPs of our management strategy.

```{r, message = FALSE, echo = TRUE}
em <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0, # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  HCR = build_hcr(HCR = 5,       # Tier3 HCR
                  Ftarget = 0.4, # F40%
                  Flimit = 0.35, # F35%
                  Plimit = 0.2,  # No fishing when SB<SB20
                  Alpha = 0.05),
  fit_control = fit_control(verbose = 1, phase = TRUE)
)
```

------------------------------------------------------------------------

## Running MSEs

```{r, message = FALSE, echo = TRUE}
mse1 <- run_mse(
  om = om, em = em,
  nsim = 10,
  assessment_period = 2,
  # Fishery = fleet 1 (annual), BTS = fleet 2 (biannual)
  sampling_period = c(1, 2),
  seed = 666 # default; per-sim seed = seed + sim
)
```

### Reproducibility

`run_mse()` accepts two seed arguments, both with default `seed = 666`:

* `seed` controls the random draws *inside* each simulated trajectory
  (recruitment, observation error, composition resampling). Each
  simulation worker calls `set.seed(seed + sim)`, so simulation `1`
  uses seed `667`, simulation `2` uses `668`, and so on. Re-running
  the same `run_mse()` call with the same `seed` reproduces the
  trajectory bit-for-bit.
* `regenerate_seed` (defaults to `seed`) controls the optional
  `regenerate_past = TRUE` step that re-fits the EM to OM-simulated
  historical data. Set this separately if you want to vary
  past-data realisations while holding the projection draws fixed.

The MSE uses a PSOCK cluster (`parallel::parLapply`) by default, so
worker processes are clean and only see the explicit objects shipped
by `clusterExport()` — there is no leakage from the parent's
`.GlobalEnv` RNG state. As a result the seed plumbing above is the
*only* thing that controls reproducibility: there is no need to call
`set.seed()` before `run_mse()` for the simulation results to be
deterministic. (You may still want `set.seed()` for any plotting code
that draws random colours or jitter.)

For management-facing runs, record `packageVersion("Rceattle")` and
the `seed` alongside the saved `.rds` so a later reviewer can rerun
the exact same trajectories from the same package tag.

------------------------------------------------------------------------

## Evaluating MSEs via plots

```{r, echo = TRUE}
plot_depletionSSB(lapply(mse1, function(x) x$OM))
```

```{r, echo = TRUE}
plot_depletionSSB(
  c(mse1$Sim_1$EM,           # EMs
    list(mse1$Sim_1$OM,      # OM
         mse1$Sim_1$OM_no_F  # OM w/ no fishing
    )),    
  line_col = c(
    paste0("grey", round(seq(80, 50, length.out = 15))),
    1, 3
  )
)
```

------------------------------------------------------------------------

## Summary statistics

```{r, warning = FALSE, echo = TRUE}
mse_summ <- mse_summary(mse1)
knitr::kable(mse_summ, digits = 2)
```

------------------------------------------------------------------------

# Alternative scenarios

------------------------------------------------------------------------

## Different BRPs

```{r, message = FALSE, echo = TRUE}
em2 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0, # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  HCR = build_hcr(HCR = 5,       # Tier3 HCR
                  Ftarget = 0.30,# F40%
                  Flimit = 0.10, # F35%
                  Plimit = 0,    # No limit
                  Alpha = 0.1),
  fit_control = fit_control(verbose = 1, phase = TRUE)
)

mse2 <- run_mse(
  om = om, em = em2, 
  nsim = 10, 
  assessment_period = 2,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## Triannual BTS

```{r, message = FALSE, echo = TRUE, eval = FALSE}
mse3 <- run_mse(
  om = om, em = em, 
  nsim = 10, 
  assessment_period = 2,
  sampling_period = c(1, 3) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## 4-year cycle and 50% sampling

```{r, message = FALSE, echo = TRUE, eval = FALSE}
mse4 <- run_mse(
  om = om, em = em, 
  nsim = 10, 
  fut_sample = 0.5,
  assessment_period = 4,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

## Climate MSE

### Climate linked OM

```{r, message = FALSE, echo = TRUE}
# Simulate random environmental series
nrdata$env_data <- data.frame(
  Year = nrdata$styr:nrdata$projyr,
  Index = 1:90 * 0.5 + rnorm(90))

# Fit OM
om2 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  recFun = build_srr(
    srr_fun = 1,    # Mean R with covariates
    srr_indices = 1),
  fit_control = fit_control(phase = TRUE)
)

plot_biomass(om2, incl_proj = TRUE)
```

### Run MSE

```{r, message = FALSE, echo = TRUE, eval = FALSE}
mse5 <- run_mse(
  om = om2, em = em, 
  nsim = 10, 
  fut_sample = 0.5,
  assessment_period = 4,
  sampling_period = c(1, 2) # Fishery = fleet 1, BTS = fleet 2
)
```

------------------------------------------------------------------------

# Overall summary

```{r, echo = TRUE}
mse_summ$MSE = 1
mse_summ2 <- mse_summary(mse2); mse_summ2$MSE = 2

knitr::kable(do.call("rbind", list(mse_summ, mse_summ2)) |>
               dplyr::relocate(MSE), digits = 2)
```
