---
title: "1. Rceattle: An Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{1. Rceattle: An Introduction}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette fits a sequence of five single- and multi-species models with `Rceattle`, applied to walleye pollock, arrowtooth flounder, and Pacific cod in the Gulf of Alaska and eastern Bering Sea. The model follows Holsman et al. (2016) and Adams et al. (2022). Every function has help available via `?`. The workflow is:

1.  Install `Rceattle`

2.  Load `Rceattle` and associated excel data

3.  Fit assessment models:

-   model 1: a single-species assessment model estimated with recruitment estimated as penalized effects;

-   model 2: three single-species assessment models jointly estimated with recruitment estimated as penalized effects;

-   model 3: as model 2, but with natural mortality estimated for each species;

-   model 4: as model 3, but with recruitment estimated as random effects;

-   model 5: full multi-species assessment models with time- and age- varying natural mortality due to predation.

4.  Compare models by AIC and Mohn's rho (retrospective analysis)

5.  Review plots of input data, diagnostics, and results.

## 1. Installation

`Rceattle` and associated dependencies and can be installed as follows:

```{r, eval=FALSE}
# Rceattle (pulls CRAN dependencies automatically; e.g. TMB, Matrix, dplyr)
install.packages("remotes")
remotes::install_github("grantdadams/Rceattle")

# Optional: TMBhelper provides richer optimization diagnostics.
# Rceattle falls back to plain nlminb + sdreport if it's not installed.
# devtools::install_github("kaskr/TMB_contrib_R/TMBhelper")
```

## 2. Data

To run `Rceattle` a data object must first be loaded. Data from the 2018 Gulf of Alaska and 2017 eastern Bering Sea single- and multi-species assessments are bundled with the package:

```{r, warning=FALSE, message=FALSE}
library(Rceattle)
data("GOApollock") # Data for GOA-pollock (1-species) ?GOApollock for more information on the data 
data("BS2017SS") # Combined data for 3 species. ?BS2017SS for more information on the data 
data("BS2017MS") # Multi-species data. Note: the only difference is the residual mortality (M1_base) is lower
```

Data can be visualized using the `plot_data` function:

```{r}
plot_data(GOApollock)
```

The data files can be written to excel, modified within excel, and read back into `R` as follows:

```{r, message = FALSE}
Rceattle::write_data(data_list = BS2017SS, file = "BS2017SS.xlsx")
mydata <- Rceattle::read_data( file = "BS2017SS.xlsx")
```

A description of the inputs and file structure can be found on the first sheet (`meta_data`) and the associated help function `?BS2017SS`.

## 3. Fit models

Models are fit with the `fit_mod` function. The first model is a single-species fit with no stock–recruit relationship to Gulf of Alaska pollock. It has 8 fleets: 1 fishery, 6 surveys, and 1 fleet turned off (its data excluded from the likelihood; see `model_1$fleet_control`).

```{r, message = FALSE}
model_1 <- Rceattle::fit_mod(data_list = GOApollock,
                            inits = NULL, # Initial parameters = 0
                            estimateMode = 0, # Estimate
                            random_rec = FALSE, # No random recruitment
                            fit_control = fit_control(
                              phase = TRUE,
                              verbose = 1))
summary(model_1)
```


The second model fits three single-species models jointly, with no stock–recruit curve and natural mortality fixed at the input values (`mydata$M1_base`). It has 7 fleets: 3 fisheries (one per species) and 4 surveys (see `mydata$fleet_control`). Single-species mode is the default (`msmMode = 0`):

```{r, message = FALSE}
model_2 <- Rceattle::fit_mod(data_list = mydata,
                            inits = NULL, # Initial parameters = 0
                            estimateMode = 0, # Estimate
                            random_rec = FALSE, # No random recruitment
                            msmMode = 0, # Single-species
                            fit_control = fit_control(
                              phase = TRUE,
                              verbose = 1))
summary(model_2)
```

Alternatively, natural mortality can be estimated with the `build_M1` function; `M1_model = "sex_age_invariant"` estimates a single M per species, constant across sex and age:

```{r, message = FALSE}
model_3 <- Rceattle::fit_mod(data_list = mydata,
                              estimateMode = 0, # Estimate
                              M1Fun = build_M1(M1_model = "sex_age_invariant"), # Estimate sex-invariant M)
                              random_rec = FALSE, # Penalized likelihood
                              msmMode = 0, # Single species mode
                              fit_control = fit_control(
                                phase = TRUE,
                                verbose = 1))
summary(model_3)
```

To estimate `model_4` where recruitment is treated as random effects we can use the `random_rec` argument. For this model we initialized at the previous model's MLEs and don't phase estimation to decrease run-time:

```{r, message = FALSE}
model_4 <- Rceattle::fit_mod(data_list = mydata,
                              inits = model_3$estimated_params, # Start estimation at model 3's MLEs
                              estimateMode = 0, # Estimate
                              M1Fun = build_M1(M1_model = "sex_age_invariant"),
                              random_rec = TRUE, # Random recruitment
                              msmMode = 0, # Single species mode
                              fit_control = fit_control(
                                phase = FALSE,
                                verbose = 1))
summary(model_4)
```

To estimate time- and age-varying natural mortality due to predation we can estimate the model in multi-species mode by setting `msmMode = 1`. This follows the MSVPA parameterization described in [Magnusson (1995)](https://link.springer.com/article/10.1007/BF00179756).

```{r, message = FALSE}
model_5 <- Rceattle::fit_mod(data_list = BS2017MS,
                            inits = model_3$estimated_params, # Initial parameters from single species estimates
                            M1Fun = build_M1(M1_model = "sex_age_invariant"),
                            estimateMode = 0, # Estimate
                            niter = 3, # 3 iterations around population and predation dynamics
                            random_rec = FALSE, # No random recruitment
                            msmMode = 1, # MSVPA based
                            suitMode = 0, # Empirical suitability
                            fit_control = fit_control(
                              verbose = 1))
summary(model_5)
```

## 4. Compare models

We can compare models by evaluating the marginal or joint negative log-likelihood, AIC, and retrospective analysis. NOTE, we can not compare models estimated using penalized likelihood vs marginal likelihood using AIC. The likelihood based components all can be found in the model objects build above:

```{r, eval = FALSE}
# Evaluate 1 model
model_1$opt$AIC # AIC
model_1$opt$objective # Negative log-likelihood
model_1$quantities$jnll_comp # Negative log-likelihood components

# Compare AIC across 3-species penalized likelihood models
sapply(list(model_2, model_3, model_5), function(x) x$opt$AIC)

```

We can also look at retrospective bias in each assessment using the `retrospective` function, which creates a list of model objects and a data.frame with mohn's rho calculations.

```{r, message = FALSE, warning = FALSE}
model_1_retro <- retrospective(model_1, peels = 7)

plot_biomass(model_1_retro$Rceattle_list, model_names = paste("Pollock; mohns =", round(model_1_retro$mohns[1,2], 3)))
```

## 5. Model plots and diagnostics

Various plots and diagnostics are built into `Rceattle`. Specifically, time-series hindcast and forecast biomass, SSB, recruitment, biomass consumed as prey, mortality-at-age, and depletion with uncertainty. The time-series plots can accept a list of models (see above).

```{r}
model_list <- list(model_2, model_3, model_4, model_5)
model_names <- list("single-spp", "Single-spp M", "Single-spp M - RE", "Multi-spp")

# Biomass
plot_biomass(model_list, model_names = model_names, incl_proj = TRUE, add_ci = TRUE)

```

```{r, eval = FALSE}
# Recruitment
plot_recruitment(model_list, model_names = model_names, incl_proj = TRUE, add_ci = TRUE)

# SSB
plot_ssb(model_list, model_names = model_names, incl_proj = TRUE, add_ci = TRUE)

# Biomass depletion
plot_depletion(model_list, model_names = model_names, incl_proj = FALSE, add_ci = FALSE)

# SSB depletion
plot_depletion(model_list, model_names = model_names, incl_proj = FALSE, add_ci = FALSE)

```

Diagnostics plots include plotting selectivity, fits to composition data, index fit, and catch fit. Selectivity and composition plot functions can only input one model. Index and catch fit plots can take multiple model inputs.

```{r, eval = FALSE}
# Selectivity
plot_selectivity(model_2)

# Plot composition fit
plot_comp(model_2)

# Plot index
plot_index(model_list) # - All models

# Plot catch data
plot_catch(model_list)
```

Outputs of the model can queried using the `$` sign:

-   `model_1$initial_params` - list of initial parameter values prior to estimation

-   `model_1$bounds` - list of parameter bounds

-   `model_1$map` - list of parameter map in factor form (`model_1$map$mapFactor`) or list form (`model_1$map$mapList`)

-   `model_1$estimated_params` - list of estimated parameters post estimation

-   `model_1$quantities` - list of derived quantities (SSB, biomass, selectivity, etc)

-   `model_1$data_list` - the input data

-   `model_1$run_time` - model run time

-   `model_1$obj` - the TMB makeadfun object.

-   `model_1$opt` - the fit_tmb optimization object.

-   `model_1$sdrep` - the TMB sdreport object.
