---
title: "3. Model diagnostics"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{3. Model diagnostics}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

Rceattle provides three layers of diagnostics: **fit plots** to visually inspect how well the model reproduces observed data, **retrospective analysis** (peels) to detect patterns of systematic over- or under-estimation, and **jitter testing** to check that the optimiser has found a global minimum.

## Setup and plotting data

```{r}
library(Rceattle)

# Fit a single-species model for the 2022 Northern Rockfish assessment.
data("NorthernRockfish2022")
nrdata <- NorthernRockfish2022
nrdata$fleet_control$proj_F_prop <- 1
plot_data(nrdata)
```

Fitting the model
```{r}
model_1 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  fit_control = fit_control(phase = TRUE, verbose = 0)
)
summary(model_1)
```

## Fit plots

### Composition data

`plot_comp()` overlays observed (bars) and predicted (line) age or length compositions for every fleet.

```{r}
plot_comp(model_1)
```

### Survey indices

`plot_index()` and `plot_logindex()` show observed vs. predicted survey biomass on the natural and log scales, respectively. `plot_indexresidual()` plots the log-scale residuals — a useful first check for time trends or heteroscedasticity.

```{r}
plot_index(model_1)
plot_logindex(model_1)
plot_indexresidual(model_1)
```

### Catch

```{r}
plot_catch(model_1)
```

## Retrospective analysis

A retrospective analysis systematically removes the most recent years of data one peel at a time and re-fits the model. Bias in the final estimates relative to earlier peels is summarised by Mohn's rho: values outside ±0.2 (for SSB) generally indicate a problem worth investigating.

```{r}
model_1_retro <- retrospective(Rceattle = model_1, peels = 5)

# Mohn's rho for each quantity
model_1_retro$mohns

# Plot historical trajectories across peels
plot_biomass(model_1_retro$Rceattle_list)

# Include the projection period to see how the forecast changes
plot_biomass(model_1_retro$Rceattle_list, incl_proj = TRUE)
```

The `retrospective()` function returns a list with two elements:

| Element | Description |
|---|---|
| `Rceattle_list` | List of fitted models, ordered from full run to most-peeled |
| `mohns` | Data frame with columns `Object` (quantity, e.g. `"Biomass"`, `"SSB"`), `Forecast year` (0 = terminal year bias; 1+ = forecast skill), `N` (number of peels), and one column per species with mean relative error (Mohn's rho) |

The `nyrs_forecast` argument (default 3) additionally evaluates rho for years projected beyond the terminal peel, making it possible to quantify forecast skill alongside retrospective bias in a single call.

## Jitter testing

Jitter testing re-fits the model from many randomly perturbed starting values to check whether the optimizer consistently returns the same minimum negative log-likelihood (NLL). A spread of NLL values across jitters suggests the likelihood surface has multiple local minima and results should be interpreted cautiously.

```{r}
jitters <- jitter(Rceattle = model_1, njitter = 10, phase = TRUE)  # default njitter = 50

# Histogram of NLL differences relative to the best run
hist(log(jitters$nll - min(jitters$nll)),
     main = "Jitter NLL spread (log scale)",
     xlab = "log(NLL - min NLL)")

# Overlay biomass trajectories — tight overlap indicates a stable optimum
plot_biomass(jitters$Rceattle_list)

# Number of runs that converged
length(jitters$Rceattle_list)
```

Non-converging runs are silently dropped from `Rceattle_list`, so if `length(jitters$Rceattle_list)` is much less than `njitter`, convergence may be fragile.

## Comparing single- and multi-species trajectories

Plotting sensitivity runs together is itself a useful diagnostic — large divergences in biomass or mortality deserve scrutiny.

```{r}
mod_list  <- list(model_1, model_1_retro$Rceattle_list$Year_2019)
mod_names <- 1:length(mod_list)

plot_biomass(Rceattle = mod_list, model_names = mod_names)
plot_recruitment(Rceattle = mod_list, model_names = mod_names, add_ci = TRUE)
plot_depletionSSB(Rceattle = mod_list, model_names = mod_names)
```


## Model average

For model averaging across model variants, see `?model_average`.

```{r}
mod_avg <- model_average(Rceattle = list(model_1, model_1), weights = c(1,1))
plot_biomass(mod_avg)
```
