---
title: "3. Model diagnostics"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{3. Model diagnostics}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

Rceattle provides a full set of model diagnostics:

- **convergence diagnostics** — attached automatically to every fit
- **S3 methods** — `residuals()`, `logLik()`, and the like
- **fit plots** — compare predicted to observed data
- **retrospective analysis** — peel off recent years to detect systematic bias
- **jitter testing** — perturb starting values to check the optimizer found a global minimum
- **self-testing** (simulation–estimation) — check the model recovers its own estimates from simulated data
- **likelihood profiling** — check how informative the data are about a given parameter

## Setup and plotting data

```{r}
library(Rceattle)

# Fit a single-species model for the 2022 Northern Rockfish assessment.
data("NorthernRockfish2022")
nrdata <- NorthernRockfish2022
nrdata$fleet_control$Proj_F_proportion <- 1
plot_data(nrdata)
```

Fit the model.
```{r}
model_1 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  fit_control = fit_control(phase = TRUE, verbose = 0)
)
summary(model_1)
```

## S3 Methods

```{r}
summary(model_1)              # same compact summary
coef(model_1)                 # estimated fixed-effect parameters
logLik(model_1)               # logLik with df attribute (AIC works)
AIC(model_1)
vcov(model_1)                 # fixed-effect covariance from sdreport
residuals(model_1)                    # response residuals, all data sources
residuals(model_1, type = "pearson")  # Pearson residuals (glm-style `type`)
as.data.frame(model_1)
```


## Convergence diagnostics

After optimization, `fit_mod()` runs a battery of convergence checks and attaches
the result as `model$convergence`. Each check is one record with a `severity`
(`"OK"`, `"NOTE"`, `"WARN"`, or `"FAIL"`); the object's `status` is the worst
severity present. Non-OK checks are reported via `message()` during the fit, and
`print(model)` shows the overall status. A non-converged model is never turned
into an error — the fit is always returned with its diagnostics attached, so you
decide whether to trust it.

```{r}
model_1$convergence            # overall status + any non-OK checks
print(model_1$convergence, all = TRUE)   # show every check, including OK ones
```

Re-run the battery on any fit with `convergence_diagnostics()`:

```{r}
convergence_diagnostics(model_1)
```

The checks cover:

- **`max_gradient`** — maximum absolute marginal gradient and the parameter
  carrying it (`WARN` > 1e-3, `FAIL` > 1).
- **`pdHess`** — whether the Hessian is positive definite (`FAIL` if not).
- **`sdreport_failed`** — `FAIL` when an `sdreport` was requested but the
  Hessian could not be inverted.
- **`hessian_conditioning`** — the Hessian condition number and, when poorly
  conditioned, the parameters loading on the least-identified direction (the
  unidentified linear combination). Complements `TMBhelper::check_estimability`
  (a per-parameter verdict) with a continuous severity and direction.
- **`parameters_on_bounds`** — parameters that hit a configured `build_bounds()`
  limit (often unidentified or mis-scaled).
- **`phasing`** — phases that ended with a high gradient, localizing which
  parameter block is hard to fit.
- **`estimability`** — surfaces `TMBhelper::check_estimability` when it ran.

## Fit plots

### Composition data

`plot_comp()` draws composition Pearson-residual bubbles and overlays the
observed (shaded area) and predicted (line) age or length compositions for every
fleet, both annually and aggregated across years. Joint-sex data are drawn on a
single bin axis with females above and males below zero.

```{r}
plot_comp(model_1)
```

### Survey indices

`plot_index()` shows observed vs. predicted survey biomass; pass `log = TRUE` for the log scale. `plot_indexresidual()` plots the log-scale residuals — a useful first check for time trends or heteroscedasticity. These return `ggplot` objects.

```{r}
plot_index(model_1)
plot_index(model_1, log = TRUE)
plot_indexresidual(model_1)
```

### Catch

```{r}
plot_catch(model_1)
```

## One-step-ahead (OSA) residuals

Pearson residuals on composition data are *not* standard normal even when the
model is correct: the age/length bins are correlated (more fish in one bin means
fewer in another), and random effects induce correlation across years. One-step-ahead
(OSA) residuals remove both problems. Each observation is residualized conditional
on the observations added before it, integrating out the random effects, so under a
correctly specified model the residuals are iid standard normal (Thygesen et al. 2017;
Trijoulet et al. 2023). This makes them suitable for objective goodness-of-fit testing.

`osa_residuals()` computes them *post hoc* for any combination of the fitted data
types -- survey indices (`"index"`), fishery catch (`"catch"`), age/length
composition (`"comp"`), conditional age-at-length (`"caal"`), predator diet
(`"diet"`), and, where a linkage carries an observed state-space covariate, the
covariate observation itself (`"ecov"`). Composition data use the conditional
binomial / beta-binomial decomposition. Because OSA re-optimizes the random
effects for each observation it is expensive, so run it on a converged fit
(`estimateMode < 3`). All sources are available from any fit; `osa_residuals()`
builds the required observation data on demand and skips a requested source the
model has no observations for (it errors only if *none* of them do). `"diet"` is
the one source not in the default set -- it applies only to multispecies models
and can be slow -- so pass `source = "all"` to include it.

```{r osa, eval = FALSE}
# Aggregate (catch + index) residuals
osa <- osa_residuals(model_1, source = c("index", "catch"))
head(osa)

# Statistical diagnostics (Stewart & Monnahan 2025): SDNR and lower/upper tail
# statistics, each with the interval expected under the standard-normal null.
osa_diagnostics(osa)

# Q-Q plot (with SDNR / tail annotation) + residual-by-year
plot(osa)

# residuals() also exposes them in the common residual schema (`source` selects
# the data, like `type` in stats::residuals.glm()):
residuals(model_1, type = "osa", source = c("index", "catch"))

# Composition OSA residuals -- no special fit required
plot_comp(model_1, residual_type = "osa")              # Q-Q + signed bubbles
residuals(model_1, type = "osa", source = "comp")
```

Following Stewart and Monnahan (2025), a practical workflow is:

1. **Inspect the aggregate fit** across years (`plot_comp(model_1)`): systematic
   lack of fit here points to structural misspecification (e.g. the wrong
   selectivity shape).
2. **Compare the SDNR** to its null interval (from `osa_diagnostics()`): a value
   outside the interval flags over- or under-dispersion, often a data-weighting
   or effective-sample-size issue.
3. **Compare the lower/upper tail statistics** to their null intervals: these
   localize departures to the tails of the residual distribution.
4. **Inspect Pearson bubble plots** (`plot_comp(model_1)`, the default): patterns
   in the sign of residuals across age/year can identify *where* lack of fit
   occurs -- Stewart and Monnahan recommend keeping these alongside OSA
   residuals.
5. **Revise the model or data weighting** where misspecification is found, using
   this together with the other diagnostics rather than as an accept/reject test.

When the diagnosis is a weighting problem rather than a structural one,
`reweight_comps()` runs the McAllister-Ianelli (1997) tuning loop: it refits the
model the way it was fitted, updating each fleet's `Comp_weights` until the
implied weights settle.

```{r}
tuned <- reweight_comps(model_1, n_iter = 10, tol = 0.01)
tuned$reweight$history    # weight per fleet, per iteration
```

The alternatives are Francis (2011) reweighting, applied by hand, and estimating
the weights directly with a Dirichlet-multinomial likelihood
(`Comp_distribution = "DirichletMultinomial"`), which avoids the iteration
entirely.

Because OSA residuals for discrete composition data are randomized quantile
residuals, they are stochastic; `osa_residuals()` takes a `seed` for
reproducibility.

### Process residuals

`process_residuals()` is the complementary check on the *process* model. It
standardizes the model's random-effect deviations (recruitment, initial
abundance, catchability) against their process prior, drawing once from the
joint posterior (SAM's `procres`; Nielsen and Berg 2014). Under a correct
process model these too are approximately iid standard normal.

```{r procres, eval = FALSE}
pr <- process_residuals(model_1, process = "recruitment")
osa_diagnostics(pr)
plot(pr)

# residuals() exposes them too:
residuals(model_1, type = "process")
```

## Retrospective analysis

A retrospective analysis removes the most recent years of data one peel at a time and re-fits the model. Mohn's rho summarizes how the terminal estimates shift across peels; for SSB, a rho outside ±0.2 generally signals a problem worth investigating.

```{r}
model_1_retro <- retrospective(Rceattle = model_1, peels = 5)

# Mohn's rho for each quantity
model_1_retro$mohns

# Plot historical trajectories across peels
plot_biomass(model_1_retro$Rceattle_list)

# Include the projection period to see how the forecast changes
plot_biomass(model_1_retro$Rceattle_list, incl_proj = TRUE)
```

Each peel carries its own terminal year as `data_list$endyr`, so the first plot draws every peel only as far as it was fit and the peels fan out.

A peel still estimates the years it dropped: they are its retrospective forecast, fit to the observed catch with recruitment held at the peel's mean and the survey and composition data withheld. Fishing mortality is estimated through the unpeeled terminal year and the harvest control rule takes over after it, so each peel carries three years:

| Field | Meaning |
|---|---|
| `endyr`, `endyr_peel` | the peel's terminal year — what it was fit through |
| `endyr_full` | the unpeeled terminal year, where the retrospective forecast ends |
| `projyr` | the end of the projection |

`incl_proj = TRUE` plots the forecast and projection years. Take the forecast years as `endyr_peel + seq_len(endyr_full - endyr_peel)` — empty for the unpeeled model, where `(endyr_peel + 1):endyr_full` would count down instead.

The `retrospective()` function returns a list with two elements:

| Element | Description |
|---|---|
| `Rceattle_list` | List of fitted models, ordered from most-peeled to full run, each named for its own terminal year (`Year_2013`, …, `Year_2017`) |
| `mohns` | Data frame with columns `Object` (quantity, e.g. `"Biomass"`, `"SSB"`), `Forecast year` (0 = terminal year bias; 1+ = forecast skill), `N` (number of peels), and one column per species with mean relative error (Mohn's rho) |

The `nyrs_forecast` argument (default 3) also evaluates rho for years projected beyond each terminal peel, so a single call quantifies forecast skill alongside retrospective bias.

A peel that did not converge is dropped and a message reports how many, so `Rceattle_list` can be shorter than `peels + 1`. Index it by name rather than position — with a peel missing, `Rceattle_list[[3]]` is not the 3-year peel. If no peel converges, Mohn's rho is `NaN` and the call warns.

## Jitter testing

Jitter testing re-fits the model from many randomly perturbed starting values and checks whether the optimizer keeps returning the same minimum negative log-likelihood (NLL). A spread of NLL values across jitters points to multiple local minima, so the results should be treated with caution.

```{r}
jitters <- jitter(Rceattle = model_1, njitter = 10, phase = TRUE)  # default njitter = 50

# Histogram of NLL differences relative to the best run
hist(log(jitters$nll - min(jitters$nll)),
     main = "Jitter NLL spread (log scale)",
     xlab = "log(NLL - min NLL)")

# Overlay biomass trajectories — tight overlap indicates a stable optimum
plot_biomass(jitters$Rceattle_list)

# Number of runs that converged
length(jitters$Rceattle_list)
```

Non-converging runs are dropped from `Rceattle_list` and reported in a message, so if `length(jitters$Rceattle_list)` is well below `njitter`, convergence is fragile. Read that count as part of the result, not as a nuisance: the dropped starts are the ones that wandered off, so a jitter test that returns only its successes understates the spread.

A run is dropped when its maximum absolute gradient exceeds 1 — the same threshold `convergence_diagnostics()` calls `FAIL` — or when an `sdreport` that was requested did not come back.

`retrospective()` and `jitter()` both fit peels / starts in parallel by default. Pass `cores = 1` to force sequential execution; otherwise the call uses a PSOCK cluster sized at `parallel::detectCores() - 6` (capped at 2 under `R CMD check`).

## Self-test (simulation–estimation)

`self_test()` simulates `nsim` datasets from a fitted model — holding its estimated parameters fixed — and re-fits the model to each. If the refit estimates cluster around the values used to simulate them, the model is at least self-consistent: it can recover its own estimates from data it generated. Persistent bias or a wide spread in a quantity means the data are not informative about it.

```{r}
sims <- self_test(model_1, nsim = 10)            # default nsim = 50

# Number of simulations that converged (non-converged runs are dropped)
length(sims)

# Overlay biomass / SSB trajectories across simulations — the original fit's
# trajectory should sit inside the spread of the refits.
plot_biomass(c(list(model_1), sims), model_names = c("fit", names(sims)))
plot_ssb(c(list(model_1), sims),     model_names = c("fit", names(sims)))
```

Each simulation uses `seed + i` as its RNG seed, so results are reproducible whether or not `cores > 1`. Set `simulate = FALSE` to refit against the model's *expected* values (no observation error) rather than draws from the observation likelihood — useful for confirming that the estimator returns the generating parameters in the noise-free limit. `cores` behaves the same way as in `retrospective()` and `jitter()`.

Every refit starts from the fitted model's *starting* values, not its estimates, so it covers the same ground the original fit did. `phase` therefore defaults to whatever the input model was fit with — pass `phase = TRUE` explicitly if you fit the model some other way and the refits are being dropped, since a model that needed phasing once needs it for every simulated data set too. `start = "estimated"` starts from the fitted estimates instead: much faster, but the optimizer then begins inside the basin containing the generating values and recovery is close to guaranteed by construction, so read it as optimistic about recovery rather than simply less powerful.

Two things to keep in mind when reading the result. Non-converged refits are dropped and reported in a message, and they are dropped *because* they behaved badly — so the returned replicates are a favourable subset, and any spread or bias computed over them (including `compare_sim()`'s CV) understates the truth. And `sim_mod()` redraws the observations only: with `random_rec = TRUE` every replicate shares the operating model's single recruitment realization, so the spread across replicates is observation error alone — a lower bound on estimation uncertainty in SSB and recruitment, not comparable to the model's own uncertainty bands.

When the returned list is shorter than `nsim`, `debug = TRUE` returns *every* simulation so the failures can be read directly rather than inferred:

```{r}
sims <- self_test(model_1, nsim = 10, debug = TRUE)

failed <- names(which(!attr(sims, "converged")))
sims[[failed[1]]]$convergence          # which check failed, and on which parameter
```

## Likelihood profile

`profile()` re-fits the model across a grid of fixed values for a chosen parameter and returns the NLL surface. A flat profile means the data carry little information about the parameter; a minimum away from the MLE means the fit has not truly settled there. It supports the recruitment standard deviation, the stock–recruit parameters, and natural mortality.

Specify the parameter with a natural-scale alias: `"sigmaR"` (recruitment
standard deviation), `"M1"` (natural mortality), or the stock–recruit parameters
`"R0"`, `"alpha"`, and `"beta"`. Aliases take natural-scale values directly.
`slots` gives the cell(s) to profile — usually just the species index — and
`values` gives the grid of values for each. Supplying more than one slot
profiles over the full grid of combinations.

```{r}
# 1-D profile: sigmaR for species 1 (natural-scale alias)
prof_sigmaR <- profile(
  fitted = model_1,
  param    = "sigmaR",
  slots    = list(1),
  values   = list(seq(0.1, 1.5, by = 0.1))
)

plot(prof_sigmaR$grid$slot_1,
     prof_sigmaR$nll - min(prof_sigmaR$nll, na.rm = TRUE),
     type = "l", xlab = "sigmaR", ylab = "dNLL")

# 1-D profile: SRR alpha for species 1
# (alias fills in the rec_pars column; slot is just the species index)
prof_alpha <- profile(
  fitted = model_1,
  param    = "alpha",
  slots    = list(1),
  values   = list(seq(2, 80, length.out = 20))
)

# 2-D cross-profile: M1 across sex for species 1, age 1
prof_M_sex <- profile(
  fitted = model_1,
  param    = "M1",
  slots    = list(c(1, 1, 1), c(1, 2, 1)),  # males and females
  values   = list(seq(0.10, 0.40, by = 0.05),
                  seq(0.10, 0.40, by = 0.05))
)
```

`profile()` returns `Rceattle_list` (one fit per grid row, `NULL` where the fit failed), `grid` (the user-scale value grid), `nll` (joint NLL aligned with `grid`, `NA` for non-converged fits), and echoes of `param` and `slots`. To cross-profile across multiple species, supply one slot per species (e.g. `slots = list(1, 2, 3)` with `param = "sigmaR"`); to cross-profile M1 across sex, supply one slot per sex as in the third example above. `cores` behaves the same way as in the other diagnostic functions.

## Comparing runs

Overlaying sensitivity runs is itself a useful diagnostic — large divergences in biomass or mortality deserve scrutiny.

```{r}
mod_list  <- list(model_1, model_1_retro$Rceattle_list$Year_2019)
mod_names <- 1:length(mod_list)

plot_biomass(Rceattle = mod_list, model_names = mod_names)
plot_recruitment(Rceattle = mod_list, model_names = mod_names, add_ci = TRUE)
plot_depletionSSB(Rceattle = mod_list, model_names = mod_names)
```


## Model average

To average estimates across model variants, see `?model_average`.

```{r}
mod_avg <- model_average(Rceattle = list(model_1, model_1), weights = c(1,1))
plot_biomass(mod_avg)
```
