---
title: "3. Model diagnostics"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{3. Model diagnostics}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

Rceattle provides several layers of diagnostics: **convergence diagnostics** attached automatically to every fit, **S3 methods** (`residual`, `logLik`, etc), **fit plots** to visually inspect how well the model reproduces observed data, **retrospective analysis** (peels) to detect patterns of systematic over- or under-estimation, **jitter testing** to check that the optimiser has found a global minimum, **self-testing** (simulation–estimation) to check that the model can recover its own estimates from simulated data, and **likelihood profiling** to check how informative the data are about a particular parameter.

## Setup and plotting data

```{r}
library(Rceattle)

# Fit a single-species model for the 2022 Northern Rockfish assessment.
data("NorthernRockfish2022")
nrdata <- NorthernRockfish2022
nrdata$fleet_control$proj_F_prop <- 1
plot_data(nrdata)
```

Fitting the model
```{r}
model_1 <- Rceattle::fit_mod(
  data_list = nrdata,
  estimateMode = 0,  # Estimate
  initMode = 1,      # Assume unfished equilibrium
  M1Fun = build_M1(
    updateM1 = TRUE,
    M1_model = 1,
    M1_use_prior = TRUE,
    M_prior = 0.06,
    M_prior_sd = 0.05),
  fit_control = fit_control(phase = TRUE, verbose = 0)
)
summary(model_1)
```

## S3 Methods

```{r}
summary(model_1)              # same compact summary
coef(model_1)                 # estimated fixed-effect parameters
logLik(model_1)               # logLik with df attribute (AIC works)
AIC(model_1)
vcov(model_1)                 # fixed-effect covariance fro
residuals(model_1)                    # response residuals, all data sources
residuals(model_1, type = "pearson")  # Pearson residuals (glm-style `type`)
as.data.frame(model_1)
```


## Convergence diagnostics

`fit_mod()` runs a battery of convergence checks after optimization and attaches
the result as `model$convergence`. Each check is one record with a `severity`
(`"OK"`, `"NOTE"`, `"WARN"`, or `"FAIL"`); the object's `status` is the worst
severity present. Non-OK checks are surfaced via `message()` during the fit (a
non-converged model is never turned into an error — the fit is always returned
with its diagnostics attached), and `print(model)` shows the overall status.

```{r}
model_1$convergence            # overall status + any non-OK checks
print(model_1$convergence, all = TRUE)   # show every check, including OK ones
```

Re-run the battery on any fit with `convergence_diagnostics()`:

```{r}
convergence_diagnostics(model_1)
```

The checks cover:

- **`max_gradient`** — maximum absolute marginal gradient and the parameter
  carrying it (`WARN` > 1e-3, `FAIL` > 1).
- **`pdHess`** — whether the Hessian is positive definite (`FAIL` if not).
- **`sdreport_failed`** — `FAIL` when an `sdreport` was requested but the
  Hessian could not be inverted.
- **`hessian_conditioning`** — the Hessian condition number and, when poorly
  conditioned, the parameters loading on the least-identified direction (the
  unidentified linear combination). Complements `TMBhelper::check_estimability`
  (a per-parameter verdict) with a continuous severity and direction.
- **`parameters_on_bounds`** — parameters that hit a configured `build_bounds()`
  limit (often unidentified or mis-scaled).
- **`phasing`** — phases that ended with a high gradient, localizing which
  parameter block is hard to fit.
- **`estimability`** — surfaces `TMBhelper::check_estimability` when it ran.

Each record carries a `data` element with the underlying numbers (e.g. the
condition number and loadings for `hessian_conditioning`).

## Fit plots

### Composition data

`plot_comp()` draws composition Pearson-residual bubbles and overlays the
observed (shaded area) and predicted (line) age or length compositions for every
fleet, both annually and aggregated across years. Joint-sex data are drawn on a
single bin axis with females above and males below zero.

```{r}
plot_comp(model_1)
```

### Survey indices

`plot_index()` and `plot_logindex()` show observed vs. predicted survey biomass on the natural and log scales, respectively. `plot_indexresidual()` plots the log-scale residuals — a useful first check for time trends or heteroscedasticity.

```{r}
plot_index(model_1)
plot_logindex(model_1)
plot_indexresidual(model_1)
```

### Catch

```{r}
plot_catch(model_1)
```

## One-step-ahead (OSA) residuals

Pearson residuals on composition data are *not* standard normal even when the
model is correct, because the age/length bins are correlated (more fish in one
bin means fewer in another) and because random effects induce correlation across
years. One-step-ahead (OSA) residuals fix this: each observation is residualized
conditional on the previously-added observations, integrating out the random
effects, so the residuals are iid standard normal under a correctly specified
model (Thygesen et al. 2017; Trijoulet et al. 2023). They therefore support
objective goodness-of-fit testing.

`osa_residuals()` computes them *post hoc* for any combination of the fitted data
types -- survey indices, fishery catch, age/length composition (`"comp"`),
conditional age-at-length (`"caal"`), and predator diet (`"diet"`). Composition
data are residualized with the conditional binomial / beta-binomial
decomposition. OSA re-optimizes the random effects per observation, so it is
expensive; run it on a converged fit (`estimateMode < 3`).

Computing OSA residuals for *composition* data (`"comp"`/`"caal"`/`"diet"`)
requires building extra observation data during the fit. That is opt-in for
speed -- it is skipped by default so repeated refits (simulation testing,
`run_mse()`) stay fast -- so fit with `fit_control(osa = TRUE)` when you intend
to use them. The aggregate index/catch OSA residuals and `process_residuals()`
work for any fit.

```{r osa, eval = FALSE}
# Aggregate (catch + index) residuals -- fast, available for any fit
osa <- osa_residuals(model_1, source = c("index", "catch"))
head(osa)

# Statistical diagnostics (Stewart & Monnahan 2025): SDNR and lower/upper tail
# statistics, each with the interval expected under the standard-normal null.
osa_diagnostics(osa)

# Q-Q plot (with SDNR / tail annotation) + residual-by-year
plot(osa)

# residuals() also exposes them in the common residual schema (`source` selects
# the data, like `type` in stats::residuals.glm()):
residuals(model_1, type = "osa", source = c("index", "catch"))

# Composition OSA residuals require the OSA data to be built at fit time:
model_1_osa <- Rceattle::fit_mod(
  data_list = nrdata, estimateMode = 0, initMode = 1,
  fit_control = fit_control(phase = TRUE, verbose = 0, osa = TRUE))
plot_comp(model_1_osa, residual_type = "osa")          # Q-Q + signed bubbles
residuals(model_1_osa, type = "osa", source = "comp")
```

Following Stewart and Monnahan (2025), a practical workflow is:

1. **Inspect the aggregate fit** across years (`plot_comp(model_1)`): systematic
   lack of fit here points to structural misspecification (e.g. the wrong
   selectivity shape).
2. **Compare the SDNR** to its null interval (from `osa_diagnostics()`): a value
   outside the interval flags over- or under-dispersion, often a data-weighting
   or effective-sample-size issue.
3. **Compare the lower/upper tail statistics** to their null intervals: these
   localize departures to the tails of the residual distribution.
4. **Inspect Pearson bubble plots** (`plot_comp(model_1)`, the default): patterns
   in the sign of residuals across age/year can identify *where* lack of fit
   occurs -- Stewart and Monnahan recommend keeping these alongside OSA
   residuals.
5. **Revise the model or data weighting** where misspecification is found, using
   this together with the other diagnostics rather than as an accept/reject test.

Because OSA residuals for discrete composition data are randomized quantile
residuals, they are stochastic; `osa_residuals()` takes a `seed` for
reproducibility.

### Process residuals

`process_residuals()` is the complementary check on the *process* model: it
standardizes the model's random-effect deviations (recruitment, initial
abundance, catchability) against their process prior, drawing once from the
joint posterior (SAM's `procres`; Nielsen and Berg 2014). Under a correct
process model these are also approximately iid standard normal.

```{r procres, eval = FALSE}
pr <- process_residuals(model_1, process = "recruitment")
osa_diagnostics(pr)
plot(pr)

# residuals() exposes them too:
residuals(model_1, type = "process")
```

## Retrospective analysis

A retrospective analysis systematically removes the most recent years of data one peel at a time and re-fits the model. Bias in the final estimates relative to earlier peels is summarised by Mohn's rho: values outside ±0.2 (for SSB) generally indicate a problem worth investigating.

```{r}
model_1_retro <- retrospective(Rceattle = model_1, peels = 5)

# Mohn's rho for each quantity
model_1_retro$mohns

# Plot historical trajectories across peels
plot_biomass(model_1_retro$Rceattle_list)

# Include the projection period to see how the forecast changes
plot_biomass(model_1_retro$Rceattle_list, incl_proj = TRUE)
```

The `retrospective()` function returns a list with two elements:

| Element | Description |
|---|---|
| `Rceattle_list` | List of fitted models, ordered from full run to most-peeled |
| `mohns` | Data frame with columns `Object` (quantity, e.g. `"Biomass"`, `"SSB"`), `Forecast year` (0 = terminal year bias; 1+ = forecast skill), `N` (number of peels), and one column per species with mean relative error (Mohn's rho) |

The `nyrs_forecast` argument (default 3) additionally evaluates rho for years projected beyond the terminal peel, making it possible to quantify forecast skill alongside retrospective bias in a single call.

## Jitter testing

Jitter testing re-fits the model from many randomly perturbed starting values to check whether the optimizer consistently returns the same minimum negative log-likelihood (NLL). A spread of NLL values across jitters suggests the likelihood surface has multiple local minima and results should be interpreted cautiously.

```{r}
jitters <- jitter(Rceattle = model_1, njitter = 10, phase = TRUE)  # default njitter = 50

# Histogram of NLL differences relative to the best run
hist(log(jitters$nll - min(jitters$nll)),
     main = "Jitter NLL spread (log scale)",
     xlab = "log(NLL - min NLL)")

# Overlay biomass trajectories — tight overlap indicates a stable optimum
plot_biomass(jitters$Rceattle_list)

# Number of runs that converged
length(jitters$Rceattle_list)
```

Non-converging runs are silently dropped from `Rceattle_list`, so if `length(jitters$Rceattle_list)` is much less than `njitter`, convergence may be fragile.

`retrospective()` and `jitter()` both fit peels / starts in parallel by default. Pass `cores = 1` to force sequential execution; otherwise the call uses a PSOCK cluster sized at `parallel::detectCores() - 6` (capped at 2 under `R CMD check`).

## Self-test (simulation–estimation)

`self_test()` simulates `nsim` datasets from a fitted model — keeping its estimated parameters fixed — and re-fits the model to each simulated dataset. If estimates from the refits cluster around the parameter values of the fit they were simulated from, the model is at least self-consistent (it can recover its own estimates from data it generated). Persistent bias or wide spread in a quantity is a sign that the data are not informative about it.

```{r}
sims <- self_test(model_1, nsim = 10)            # default nsim = 50

# Number of simulations that converged (non-converged runs are dropped)
length(sims)

# Overlay biomass / SSB trajectories across simulations — the original fit's
# trajectory should sit inside the spread of the refits.
plot_biomass(c(list(model_1), sims), model_names = c("fit", names(sims)))
plot_ssb(c(list(model_1), sims),     model_names = c("fit", names(sims)))
```

Each simulation uses `seed + i` as its RNG seed, so results are reproducible whether or not `cores > 1`. Set `simulate = FALSE` to refit against the model's *expected* values (no observation error) rather than draws from the observation likelihood — useful for confirming that the estimator returns the generating parameters in the noise-free limit. `cores` behaves the same way as in `retrospective()` and `jitter()`.

## Likelihood profile

`profile()` re-fits the model across a grid of fixed values for one or more cells of a chosen parameter and returns the resulting NLL surface. A flat profile means the data carry little information about the parameter; a sharp minimum away from the MLE means the fit has not actually settled there. The function generalises the legacy `profile_rsigma()` helper from `examples/Profiling_sigmaR.R` and works for any parameter slot in `Rceattle$estimated_params`; it is tested for `R_log_sd` (recruitment sigma), `rec_pars` (stock-recruit parameters), and `log_M1` (residual natural mortality).

Two ways to specify `param`:

- **Natural-scale alias** (recommended for the three documented parameters). The alias implies `transform = "log"` so you pass natural-scale values like sigmaR or M directly. For the `rec_pars` aliases (`"R0"`, `"alpha"`, `"beta"`) the column is filled in from the alias name, so `slots` only needs the species index.
  - `"sigmaR"` / `"R_sd"` &rarr; `R_log_sd`
  - `"M1"` &rarr; `log_M1`
  - `"R0"` &rarr; `rec_pars[, 1]`
  - `"alpha"` &rarr; `rec_pars[, 2]`
  - `"beta"` &rarr; `rec_pars[, 3]`
- **Raw parameter slot**. Pass the storage name (`"R_log_sd"`, `"rec_pars"`, `"log_M1"`, etc.); `slots` must index the full array and `transform` controls the scale.

`slots` is a list of integer index vectors — one entry per cell to hold fixed — and (under raw names) its entries must match the dimensionality of the parameter array (length 1 for vectors, length 2 for matrices, length 3 for 3-D arrays). `values` is a list of value grids parallel to `slots`; the full grid of fits is `expand.grid(values)`, so multiple slots produce an N-D cross-profile.

```{r}
# 1-D profile: sigmaR for species 1 (natural-scale alias)
prof_sigmaR <- profile(
  fitted = model_1,
  param    = "sigmaR",
  slots    = list(1),
  values   = list(seq(0.1, 1.5, by = 0.1))
)

plot(prof_sigmaR$grid$slot_1,
     prof_sigmaR$nll - min(prof_sigmaR$nll, na.rm = TRUE),
     type = "l", xlab = "sigmaR", ylab = "dNLL")

# 1-D profile: SRR alpha for species 1
# (alias fills in the rec_pars column; slot is just the species index)
prof_alpha <- profile(
  fitted = model_1,
  param    = "alpha",
  slots    = list(1),
  values   = list(seq(2, 80, length.out = 20))
)

# 2-D cross-profile: M1 across sex for species 1, age 1
prof_M_sex <- profile(
  fitted = model_1,
  param    = "M1",
  slots    = list(c(1, 1, 1), c(1, 2, 1)),  # males and females
  values   = list(seq(0.10, 0.40, by = 0.05),
                  seq(0.10, 0.40, by = 0.05))
)
```

`profile()` returns `Rceattle_list` (one fit per grid row, `NULL` where the fit failed), `grid` (the user-scale value grid), `nll` (joint NLL aligned with `grid`, `NA` for non-converged fits), and echoes of `param` and `slots`. To cross-profile across multiple species, supply one slot per species (e.g. `slots = list(1, 2, 3)` with `param = "sigmaR"`); to cross-profile M1 across sex, supply one slot per sex as in the third example above. `cores` behaves the same way as in the other diagnostic functions.

## Comparing single- and multi-species trajectories

Plotting sensitivity runs together is itself a useful diagnostic — large divergences in biomass or mortality deserve scrutiny.

```{r}
mod_list  <- list(model_1, model_1_retro$Rceattle_list$Year_2019)
mod_names <- 1:length(mod_list)

plot_biomass(Rceattle = mod_list, model_names = mod_names)
plot_recruitment(Rceattle = mod_list, model_names = mod_names, add_ci = TRUE)
plot_depletionSSB(Rceattle = mod_list, model_names = mod_names)
```


## Model average

For model averaging across model variants, see `?model_average`.

```{r}
mod_avg <- model_average(Rceattle = list(model_1, model_1), weights = c(1,1))
plot_biomass(mod_avg)
```
