---
title: "An overview of Rceattle"
author: "Grant Adams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{0. Rceattle: Overview}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## `Rceattle` overview

Flexible code-base for single- and multi-species age structured modelling

Empirical weight-at-age (internal growth estimation in dev)

Species-, sex- and -age structured

Flexible in the number of species and fleets

Forecasting, simulation, and MSE capable

## Background

Developed from Kirstin's ADMB CEATTLE

Ported to TMB during my PhD

Code base and functionality developed to closely/exactly match CEATTLE (ADMB), GOA Pollock (ADMB/TMB), GOA ATF (ADMB), GOA Cod & Pacific Hake (Stock Synthesis)

Designed to fit different model formulations, include environmental indices/predation, and run MSEs

## Key links

GitHub: <https://github.com/grantdadams/Rceattle>

[On-boarding doc](https://docs.google.com/document/d/1JQiyiWdWmozMMWAuObB8Eo_u09peQBFUbJFBcwP51m8/edit?usp=sharing)

## Existing `Rceattle` applications

-   Pacific hake: 1 species (Sophia Wasserman)

-   GOA ATF SAFE model: 1 species (Kalei Shotwell and Grant Adams)

-   Hake + ATF: 2 species (Raquel Ruiz Diaz)

-   EBS CEATTLE: 3 species (Kirstin Holsman)

-   GOA CEATTLE: 4 species (Grant Adams)

## Motivation and goals

1.  Tool for "easily" fitting single- and multi-species model variants

2.  Fit models with random effects

3.  Easily assess model performance via diagnostics, fit metrics, simulation, and MSE

4.  Help build environmentally-linked research models and MSE test SAFE models

5.  Eye towards flexibility and documentation from PhD angst recreating custom/hard-coded models

## Broad `Rceattle` workflow

::: {.column width="50%"}
Excel document to specify observation model:

-   Data (age/length composition, indices, catch)

-   Observation model specifications (q, selectivity, likelihood)

-   "Fixed" inputs (weight-at-age, M, maturity, sex ratio, transition matrix, ageing error)
:::

::: {.column width="50%"}
`R` wrapper function `fit_mod` to specify and fit

-   Population dynamics specifications (M, initial conditions, recruitment, predation)

-   Forecast (HCR form, BRPs, recruitment)

-   Estimation controls
:::

------------------------------------------------------------------------

## Broad `Rceattle` workflow

```{r, eval = FALSE, echo = TRUE}
library(Rceattle)

mydata <- read_data( file = "mydata.xlsx") # Reads in as list of data.frames

mymodel <- fit_mod(data_list = mydata,
                   inits = NULL,       # Initial parameters at default
                   estimateMode = 0,   # Estimate
                   random_rec = FALSE, # No random recruitment
                   msmMode = 0,        # Single species mode
                   phase = TRUE,       # Phase model
                   verbose = 1)        # Minimal messaging

plot_biomass(mymodel)
```

## Outputs

```{r, echo = TRUE}
library(Rceattle)

data("GOApollock")                              # 2018 Pollock model 

mymodel <- Rceattle::fit_mod(data_list = GOApollock,
                            inits = NULL,       # Initial parameters = 0
                            file = NULL,        # Don't save
                            estimateMode = 0,   # Estimate
                            random_rec = FALSE, # No random recruitment
                            msmMode = 0,        # Single species mode
                            phase = TRUE,
                            verbose = 1)
```

```{r, echo = TRUE}
names(mymodel)
```

## Basic excel data

-   meta_data: overview and description

-   control: species, year, age, length

-   fleet_control: observation model specification

-   index_data: survey or fishery indices (biomass/numbers)

-   catch_data: fishery catches

-   comp_data: age/length comp

-   weight: empirical weight-at-age (time-varying, fleet specific)

-   maturity: maturity-at-age (time-invariant)

-   sex_ratio: sex-ratio-at-age (time-invariant)

## Selectivity specifications (`fleet_control`)

::::: columns
::: {.column width="50%"}
`Selectivity`

-   0: empirical selectivity
-   1: logistic (2-parameter: slope and inflection)
-   2: non-parametric (Ianelli/AMAK; `Sel_curve_pen1` = monotonicity penalty, `Sel_curve_pen2` = curvature penalty)
-   3: double logistic (Dorn and Methot 1990)
-   4: descending logistic
-   5: non-parametric (Taylor/Hake cumulative coefficients)
-   6: 2D AR1 (age × year random effects)
-   7: 3D AR1 / GMRF (sparse precision matrix)

`flt_sel_dim`

-   0: age-based selectivity
-   1: length-based (converted to age via growth transition matrix)
:::

::: {.column width="50%"}
`Time_varying_sel`

-   0: no
-   1: penalized deviates (fixed `Time_varying_sel_sd_prior` as sd; requires `random_sel = FALSE` in `fit_mod`)
-   1: penalized deviates with estimated sd ($\hat{\sigma}$ estimated; requires `random_sel = TRUE` in `fit_mod` for Laplace)
-   3: time blocks (R-side mapping; no likelihood term in TMB)
-   4: random walk (penalizes year-to-year differences)
-   5: random walk on ascending portion of double logistic only.
:::
:::::

:::

## Catchability specifications (`fleet_control`)

::::: columns
::: {.column width="50%"}
`Catchability`

-   0 = Fixed at `Q_prior`
-   1 = Free parameter
-   2 = Free w/ lognormal prior with median `Q_prior` and sd `Q_sd_prior`
-   3 = Analytical (Ludwig and Walters/Martell 1994)
-   4 = Power equation (declared but currently non-functional)
-   5 = $\mu_q + X * \beta$ (environmental linkage)
-   6 = sensu Rogers et al (2024; AR1 deviates fit to environmental index)
:::

::: {.column width="50%"}
`Time_varying_q`

-   0: no
-   1: penalized deviates (IID normal, fixed `Time_varying_q_sd_prior` as sd; requires `random_q = FALSE` in `fit_mod`)
-   1: penalized deviates with estimated sd (IID normal, initialized at `Time_varying_q_sd_prior` as sd; requires `random_q = TRUE` in `fit_mod` for Laplace)
-   3: time blocks (R-side mapping; no likelihood term in TMB)
-   4: random walk (penalizes year-to-year differences)
:::
:::::

------------------------------------------------------------------------

## Recruitment specifications

::::: columns
::: {.column width="50%"}
`srr_fun` (hindcast) / `srr_pred_fun` (projections/BRPs)

-   0: mean recruitment ($R = R_0 \cdot e^{\epsilon}$)
-   1: mean recruitment with environmental linkage
-   2: Beverton-Holt ($R = \frac{\alpha \cdot SSB}{1 + \beta \cdot SSB} \cdot e^{\epsilon}$)
-   3: Beverton-Holt with environmental effect on $\alpha$
-   4: Ricker ($R = \alpha \cdot SSB \cdot e^{-\beta \cdot SSB} \cdot e^{\epsilon}$)
-   5: Ricker with environmental effect on $\alpha$

When `srr_fun = 0` and `srr_pred_fun > 0`, an Ianelli-style SRR penalty informs projections without fitting the SRR in the hindcast.

`srr_est_mode`

-   0: fix $\alpha$ to prior mean (not estimated)
-   1: freely estimate $\alpha$ and $\beta$ (no prior)
-   2: lognormal prior on steepness (B-H) or $\alpha$ (Ricker)
-   3: beta-distributed prior on steepness (B-H only)
:::

::: {.column width="50%"}
```{r, eval = FALSE, echo = TRUE}
?build_srr # See options

mymodel <- fit_mod(
  recFun = build_srr(
    srr_fun = 3,      # Beverton-Holt
    srr_pred_fun = 3, # B-H for projections
    proj_mean_rec = FALSE,
    srr_est_mode = 3, # Beta steepness prior
    srr_prior = 0.8,
    srr_prior_sd = 0.1,
    srr_indices = NA, # No environmental indices
    Bmsy_lim = NA),
  ...
)
```
:::
:::::

------------------------------------------------------------------------

## Mortality specifications

::::: columns
::: {.column width="50%"}
`M1_model`

-   0: fixed input M-at-age (not estimated)
-   1: estimate single species-specific, sex- and age-invariant M
-   2: estimate sex-specific, age-invariant M
-   3: estimate sex- and age-specific M
-   4: environmentally driven, sex-invariant M ($M_{spp,yr} = M_{spp} \cdot e^{X \cdot \beta}$)
-   5: environmentally driven, sex-specific M

`M1_re` (random effects on M)

-   0: no random effects
-   1: IID by age, constant over years
-   2: IID by year, constant over ages
-   3: IID by age and year
-   4: AR1 by age, constant over years
-   5: AR1 by year, constant over ages
-   6: 2D AR1 (separable AR1 × AR1 by age and year)

Optional lognormal prior via `M1_use_prior`, `M_prior`, `M_prior_sd`.
:::

::: {.column width="50%"}
```{r, eval = FALSE, echo = TRUE}
?build_M1 # See options

mymodel <- fit_mod(
  M1Fun = build_M1(
    M1_model = 2, # Sex-specific M
    M1_re = 6,    # 2D-AR1 on age/year
    M1_use_prior = FALSE,
    M_prior = 0.30,
    M_prior_sd = 0.05,
    M1_indices = NA),
  ...
)
```
:::
:::::

------------------------------------------------------------------------

## Harvest control rules

`HCR`

-   0: no fishing ($F = 0$)
-   1: CMSY (maximize catch subject to depletion constraint)
-   2: constant F (set to `Ftarget`)
-   3: constant F to achieve X% of $SSB_0$ (set via `Ptarget`)
-   4: constant $F_{SPR}$ with multiplier (`Ftarget * Fmult`)
-   5: NPFMC Tier 3 (linear ramp with `Alpha` and `Plimit`)
-   6: PFMC Category 1 (with `Ptarget`, `Plimit`, `Flimit`, and `Pstar`)
-   7: SESSF Tier 1 (linear ramp toward `Ftarget`)

`DynamicHCR`: FALSE = equilibrium BRPs; TRUE = dynamic BRPs

::::: columns
::: {.column width="65%"}
```{r, eval = FALSE, echo = TRUE}
?build_hcr # See options

mymodel <- fit_mod(
  HCR = build_hcr(
    HCR = 5,        # Tier 3 HCR
    DynamicHCR = FALSE, # Equilibrium BRPs
    Ftarget = 0.40, # Target F-spr
    Flimit = 0.35,  # Limit F-spr
    Ptarget = NA,
    Plimit = 0.2,   # No F < 20% SB0
    Alpha = 0.05,
    Pstar = NA,
    Sigma = NA,
    Fmult = NA,
    HCRorder = NA),
  ...
)
```
:::

::: {.column width="35%"}
```{r, echo = FALSE}
spr_fun <- function(F){
  ages = 1:10
  wt = 1:10
  sel = 1/(1+exp(-1*ages + 2))
  mat = 1/(1+exp(-1*ages + 6))
  M1 <- -0.2 * ages
  spr0 = sum( exp(M1) * wt * mat * 0.5)
  sprF = sum( exp(M1 - sel*F) * wt * mat * 0.5)
  return(sprF/spr0)
}
F30 = uniroot(function(x) spr_fun(x) - 0.30, c(0,10))$root
F35 = uniroot(function(x) spr_fun(x) - 0.35, c(0,10))$root
F40 = uniroot(function(x) spr_fun(x) - 0.4, c(0,10))$root
F48 = uniroot(function(x) spr_fun(x) - 0.48, c(0,10))$root
F20 = uniroot(function(x) spr_fun(x) - 0.2, c(0,10))$root
F45 = uniroot(function(x) spr_fun(x) - 0.45, c(0,10))$root

tier3hcr_fun <- function(Pcur = 0.5, Ptarget = 0.4, Plimit = 0, Alpha = 0.05, Ftarget = F40, Flimit = F35){
  if(Pcur > Ptarget){
   Ftarget = Ftarget
   Flimit = Flimit
  } 
  if(Alpha< Pcur/Ptarget & Pcur/Ptarget <= 1){
    Flimit = Flimit * (Pcur/Ptarget - Alpha)/(1-Alpha)
    Ftarget = Ftarget * (Pcur/Ptarget - Alpha)/(1-Alpha)
  }
  if(Alpha>=Pcur/Ptarget ){
    Flimit = 0
    Ftarget = 0
  }
  if(Pcur < Plimit ){
    Flimit = 0
    Ftarget = 0
  }
  return(list(Flimit = Flimit, Ftarget = Ftarget))
}


sessfcat1hcr_fun <- function(Pcur = 0.5, Ptarget = 0.35, Plimit = 0.2, Ftarget = F48, Flimit = F20){
  if(Pcur > Ptarget){
    Ftarget = Ftarget
  } 
  if(Pcur < Ptarget & Pcur> Plimit){
    Ftarget = Ftarget * (Pcur/Plimit - 1)
  }
  if(Pcur < Plimit ){
    Ftarget = 0
  }
  return(list(Flimit = Flimit, Ftarget = Ftarget))
}

pfmccat1hcr_fun <- function(Pcur = 0.5, Ptarget = 0.40, Plimit = 0.1, Flimit = F45, sigma = 0.5, Pstar = 0.45){
  if(Pcur > Ptarget){
    Ftarget = qnorm(Pstar, Flimit, sigma)
  } 
  if(Pcur < Ptarget & Pcur> Plimit){
    Ftarget = qnorm(Pstar, Flimit, sigma) * Ptarget * (Pcur - Plimit)/((Ptarget - Plimit)*Pcur)
  }
  if(Pcur < Plimit ){
    Ftarget = 0
  }
  return(list(Flimit = Flimit, Ftarget = Ftarget))
}


Pvals <- seq(0,1,length.out = 1000)
Tier3Ftargets <- sapply(Pvals, function(x) tier3hcr_fun(Pcur = x, Plimit = 0.2, Alpha = 0.05)$Ftarget)
Tier3Flimits <- sapply(Pvals, function(x) tier3hcr_fun(Pcur = x, Plimit = 0.2, Alpha = 0.05)$Flimit)

Tier3FtargetsATF <- sapply(Pvals, function(x) tier3hcr_fun(Pcur = x, Plimit = 0, Alpha = 0.05)$Ftarget)
Tier3FlimitsATF <- sapply(Pvals, function(x) tier3hcr_fun(Pcur = x, Plimit = 0, Alpha = 0.05)$Flimit)

SESSFCat1Ftargets <- sapply(Pvals, function(x) sessfcat1hcr_fun(Pcur = x)$Ftarget)
SESSFCat1Flimits <- sapply(Pvals, function(x) sessfcat1hcr_fun(Pcur = x)$Flimit)

PFMCCat1Ftargets <- sapply(Pvals, function(x) pfmccat1hcr_fun(Pcur = x)$Ftarget)
PFMCCat1Flimits <- sapply(Pvals, function(x) pfmccat1hcr_fun(Pcur = x)$Flimit)

PFMCCat1FtargetsFlatfish <- sapply(Pvals, function(x) pfmccat1hcr_fun(Pcur = x, Ptarget = 0.25, Plimit = 0.05, Flimit = F30, sigma = 0.5, Pstar = 0.45)$Ftarget)


#--------------------------
# Plot it
par(mar = c(3, 3, 0.5, 0.5))

# Colors (by EM/HCR)
MPcols <- 1:8 # gmRi::gmri_pal("mixed")(10)[1:8]

plot(NA,NA, ylim = c(0, 1.4), xlim = c(0,1), ylab = NA, yaxt = "na", xlab = "Depletion")
mtext(side = 2, "Fishing mortality (F)", line = 1.5)
mtext(side = 1, "Relative depletion", line = 2)

# abline(h = Tier3Ftargets[length(Tier3Ftargets)], lwd = 3, col = SamplingCol[4])

# - Tier 3
lines(x = Pvals, y = Tier3Ftargets, col = MPcols[1], lwd = 3, lty = 2)
lines(x = Pvals, y = Tier3FtargetsATF, col = MPcols[1], lwd = 3, lty = 1)
text(x=Pvals[1000], y = Tier3Ftargets[1000]+0.07, expression('1) F'["40%"]), offset = -2,pos = 4, col = MPcols[1])
#lines(x = Pvals, y = Tier3Flimits, col = MPcols[1], lwd = 3, lty = 2)


# - Cat 1
# -- 40-10
lines(x = Pvals, y = PFMCCat1Ftargets, col = MPcols[3], lwd = 3, lty = 1)
text(x=Pvals[900], y = PFMCCat1Ftargets[1000]+0.07, expression(paste("2a) ", Phi,'(0.45, F'["45%"],", 0.5)")), offset = -6,pos = 4, col = MPcols[3])
#lines(x = Pvals, y = PFMCCat1Flimits, col = 1, lwd = 3, lty = 2)

# -- 25-5
lines(x = Pvals, y = PFMCCat1FtargetsFlatfish, col = MPcols[4], lwd = 3, lty = 1)
text(x=Pvals[900], y = PFMCCat1FtargetsFlatfish[1000]+0.07, expression(paste("2b) ", Phi,'(0.45, F'["30%"],", 0.5)")), offset = -6,pos = 4, col = MPcols[4])
#lines(x = Pvals, y = PFMCCat1Flimits, col = MPcols[3], lwd = 3, lty = 2)


# - Tier 1
lines(x = Pvals, y = SESSFCat1Ftargets, col = MPcols[5], lwd = 3, lty = 1)
text(x=0.4, y = SESSFCat1Ftargets[1000]+0.07, expression('3) F'["48%"]), offset = 0,pos = 4, col = MPcols[5])


# - F40
abline(h = 0.75 * F40, col = MPcols[7], lwd = 3)
text(x=Pvals[1000], y = 0.75 * F40-0.07, expression('4) 0.75 * F'["40%"]), offset = -6 ,pos = 4, col = MPcols[7])


legend("bottomright", c("NPFMC", "PFMC 40-10", "PFMC 25-5", "SESSF", "NEFMC (Fspr)"), col = c(MPcols[1], MPcols[3], MPcols[4], MPcols[5], MPcols[7]), lty = 1, lwd = 4, bty = "n")
```
:::
:::::

------------------------------------------------------------------------

## Multi-species mode (`msmMode`)

-   0: single species (no predation mortality)
-   1: Type II MSVPA (sensu Holsman et al. 2015)
-   2: Type III MSVPA

`suitMode` (predator-prey suitability, per predator species)

-   0: empirical (MSVPA-style from diet data)
-   2: gamma selectivity on weight
-   4: lognormal on weight
-   5: normal on length
-   6: normal on weight

------------------------------------------------------------------------

## Initial conditions (`initMode`)

-   0: free parameters for initial N-at-age
-   1: equilibrium, $F_{init} = 0$, no init devs (unfished)
-   2: equilibrium, $F_{init} = 0$, with init devs (default)
-   3: non-equilibrium: $F_{init}$ estimated, init devs included
-   4: non-equilibrium: $F_{init}$ scales $R_0$ (alternative plus-group treatment)

------------------------------------------------------------------------

## Observation model specifications (`fleet_control`)

`Comp_loglike` (composition data likelihood)

-   -1: AFSC-style multinomial ($\sum n \cdot \hat{p} \cdot \log(\hat{p}/p)$)
-   0: full multinomial
-   1: Dirichlet-multinomial

`Estimate_index_sd` (survey index variance)

-   0: use provided standard deviation from data
-   1: estimate standard deviation
-   2: analytical sigma (Ludwig and Walters 1994)

------------------------------------------------------------------------

## Time-series plotting

::::: columns
::: {.column width="50%"}
```{r, eval = FALSE, echo = TRUE}
plot_biomass(mymodel)
plot_recruitment(list(mymodel1, mymodel2))
plot_depletion()
plot_depletionSSB()
plot_mortality()
```
:::

::: {.column width="50%"}
```{r, echo = FALSE}
data("BS2017SS")
BS2017SS$fleet_control$proj_F_prop <- 1
mymodel <- Rceattle::fit_mod(data_list = BS2017SS,
                            inits = NULL,       # Initial parameters = 0
                            file = NULL,        # Don't save
                            estimateMode = 0,   # Estimate
                            random_rec = FALSE, # No random recruitment
                            msmMode = 0,        # Single species mode
                            phase = TRUE,
                            verbose = 1)

mymodel_tier3 <- Rceattle::fit_mod(data_list = BS2017SS,
                            inits = NULL, # Initial parameters = 0
                            file = NULL, # Don't save
                            estimateMode = 0, # Estimate
                            random_rec = FALSE, # No random recruitment
                                  HCR = build_hcr(HCR = 5, # Tier3 HCR
                                                  Ftarget = 0.4, # F40%
                                                  Flimit = 0.35, # F35%
                                                  Plimit = 0.2, # No fishing when SB<SB20
                                                  Alpha = 0.05),
                            msmMode = 0, # Single species mode
                            phase = TRUE,
                            projection_uncertainty = TRUE,
                            verbose = 1)


# Single-species, but estimate M
ss_run_M <- Rceattle::fit_mod(data_list = BS2017SS,
                              inits = mymodel$estimated_params, # Initial parameters = 0
                              file = NULL, # Don't save
                              estimateMode = 0, # Estimate
                              M1Fun = build_M1(M1_model = 1,
                                               M1_use_prior = FALSE,
                                               M2_use_prior = FALSE),
                              random_rec = FALSE, # No random recruitment
                              msmMode = 0, # Single species mode
                              phase = TRUE,
                              verbose = 1)

ss_run_M_tier3 <- Rceattle::fit_mod(data_list = BS2017SS,
                              inits = mymodel$estimated_params, # Initial parameters = 0
                              file = NULL, # Don't save
                              estimateMode = 0, # Estimate
                              M1Fun = build_M1(M1_model = 1,
                                               M1_use_prior = FALSE,
                                               M2_use_prior = FALSE),
                                  HCR = build_hcr(HCR = 5, # Tier3 HCR
                                                  Ftarget = 0.4, # F40%
                                                  Flimit = 0.35, # F35%
                                                  Plimit = 0.2, # No fishing when SB<SB20
                                                  Alpha = 0.05),
                              random_rec = FALSE, # No random recruitment
                              msmMode = 0, # Single species mode
                              phase = TRUE,
                              projection_uncertainty = TRUE,
                              verbose = 1)

plot_biomass(list(mymodel, mymodel_tier3, ss_run_M, ss_run_M_tier3), 
             model_names = c("Fixed M", "- Tier 3", "Est M", "- Tier 3"), 
             add_ci = TRUE, incl_proj = T, species = 1)
```
:::
:::::

## Diagnostics: `??Rceattle`

```{r, echo = TRUE}
plot_index(mymodel, species = 1)
```

------------------------------------------------------------------------

```{r, echo = TRUE, eval = FALSE}
plot_comp(mymodel)
```

## Retrospectives

```{r, echo = TRUE}
retro <- retrospective(mymodel, peels = 6)
plot_biomass(retro$Rceattle_list, species = 1)
```

## MSE

```{r, echo = TRUE}
# ?run_mse for more options

mse <- run_mse(
  om = mymodel, 
  em = mymodel_tier3, 
  nsim = 1, 
  simulate_data = TRUE, 
  sample_rec = FALSE)
```

------------------------------------------------------------------------

::::: columns
::: {.column width="50%"}
```{r, echo = TRUE}
plot_depletionSSB(mse$Sim_1$OM, species = 1, model_names = "OM")
```
:::

::: {.column width="50%"}
```{r, echo = TRUE}
plot_depletionSSB(mse$Sim_1$EM, species = 1, model_names = "EMs")
```
:::
:::::

## MCMC with `tmbstan`

```{r, echo = TRUE, eval = FALSE}
## Run in parallel
library(tmbstan)

warmup = 6000       
thin = 1
iter = 6000    

cores <- 3
options(mc.cores = cores)
fit <- tmbstan(mymodel$obj, 
               init = mymodel$opt$par, # Start from MLE
               chains=cores, 
               open_progress=FALSE,
               warmup = warmup,      
               thin = thin,
               iter = iter * thin + warmup,     
               control = list(max_treedepth = 14))

## To explore the fit use shinystan
library(shinystan)
launch_shinystan(fit)
```
